---
title: "intertidal_analysis_short"
author: "tena"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
  github_document:
    html_preview: false
params:
  keep_times: ["Dawn", "Dusk"]
  control_label: "Surface Water"
---

# Overview

This document:

- Loads and merges pH/TA/nutrients and fDOM datasets
- Cleans data and applies outlier handling (IQR rule)
- Summarizes by habitat and diel period
- Creates PCA for seasonal drivers
- Plots relationships between variables to visualize processes
- Visualizes diel and habitat patterns (boxplots)

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libs}
library(tidyverse)
library(patchwork) # combine plots
library(rstatix)
library(ggcorrplot)
library(ggrepel)
library(here) # for collaboration across folders
library(lubridate)
library(kableExtra)
library(car)
library(emmeans) # for pairwise comparisons
library(agricolae)
library(lme4) # random effects
library(paletteer) # color palettes
library(ggh4x)
library(mgcv)
library(performance) 
library(MuMIn)       
library(rsample)     
library(yardstick)   
library(gratia) 
library(ggnewscale) # multiple color scales on one plot
library(vegan) # for permanova
library(lmerTest) # random effects
library(gamm4)
library(maps) # for CA map

# Global plot theme
theme_set(theme_bw(base_size = 12))
```

## Paths & Parameters

```{r params}
keep_times <- params$keep_times
control_label <- params$control_label
```

# Data Loading & Merging

This section loads both fDOM and pH/nutrients CSVs, parses dates, and joins on shared identifiers. We keep all samples, regardless of missing data (`full_join`).

```{r data_load}
load_ok <- TRUE
tryCatch({
  data_ph <- read_csv(here::here("Data","Biogeochemistry","pHProbe_Data_calculated.csv"), na = c("NA", ""), show_col_types = FALSE) 
    #mutate(Sampling_Date = lubridate::mdy(Sampling_Date))
  data_fdom <- read_csv(here::here("Data","Biogeochemistry","fDOM_clean.csv"), na = c("NA", ""), show_col_types = FALSE) %>%
    mutate(Sampling_Date = lubridate::mdy(Sampling_Date))
  data_tidal <- read.csv(here::here("Data","tidal_exchange.csv")) %>%
    mutate(Sampling_Date = lubridate::mdy(Sampling_Date))

  data_merged <- full_join(
    data_ph,
    data_fdom,
    by = c("UniqueID", "Benthos", "Quad_ID", "Day_Night", "Sampling_Date", "Sampling_Time","Season")
  ) %>%
    mutate(Humic = UVHumic+VisibleHumic+MarineHumic,# make a column for humics
           Prot = Tryptophan+Tyrosine, # make a column for proteinaceaus fDOM
           NOx = NO3+NO2 # make a column of total N
           )
  data_merged <- data_merged %>%
    left_join(data_tidal, by = c("Sampling_Date", "Day_Night"))
  
  cat("Data loaded and merged successfully.\n")
  cat("Total merged samples:", nrow(data_merged), "\n")
}, error = function(e){
  load_ok <<- FALSE
  message("Error loading or merging files. Please check file paths and names.")
  message(e)
})

sessile <- read_csv(here("Data","Community_Composition","sessile_data.csv"))
sessile_function <-read_csv(here("Data","Community_Composition","SessileFunctional.csv"))

mobile <- read_csv(here("Data","Community_Composition","mobile_data.csv"))
mobile_function <-read_csv(here("Data","Community_Composition","MobileFunctional.csv"))

data_full <- read_csv(here("Data", "Biogeochemistry", "FullDataSet_raw.csv"))

# bring in HOBO data, need to trim to experimental dates
temperature <- bind_rows(
  read_csv(here("Data","Temperature","Spring_2024", "detided", "TempLight_Spring.csv")) %>%
    filter(datetime >= as.Date("2024-04-03") & 
            datetime <= as.Date("2024-05-05")) %>%
    mutate(Season = "Spring"),
  read_csv(here("Data","Temperature","Fall_2023", "detided", "TempLight_Fall.csv")) %>%
    filter(datetime >= as.Date("2023-10-21") & 
            datetime <= as.Date("2023-11-03")) %>%
    mutate(Season = "Fall")
) %>%
   mutate(
      Benthos = case_when(
      str_detect(plot, "^B")  ~ "Barnacle",
      str_detect(plot, "^M")  ~ "Mussel",
      str_detect(plot, "^P")  ~ "Surfgrass",
      str_detect(plot, "^RW") ~ "Rockweed",
      TRUE ~ NA_character_
      )
    )

cuti <- bind_rows(read_csv(here("Data","CUTI_daily.csv")) %>%
  mutate(Date = make_date(year, month, day)) %>%
  filter(Date >= as.Date("2024-04-03") & 
            Date <= as.Date("2024-05-05")) %>%
  mutate(Season = "Spring") %>%
  select("Date","34N", "Season"),
  read_csv(here("Data","CUTI_daily.csv")) %>%
  mutate(Date = make_date(year, month, day)) %>%
  filter(Date >= as.Date("2023-10-21") & 
            Date <= as.Date("2023-11-03")) %>%
  mutate(Season = "Fall")%>%
  select("Date","34N", "Season")) %>%
  rename("index" = "34N")

temp <- bind_rows(read_csv(here("Data","wave_height_2024.csv")) %>%
            mutate(Date = make_datetime(YY, MM, DD, hh, mm)) %>%
            filter(Date >= as.Date("2024-04-03") & 
            Date <= as.Date("2024-05-05")) %>%
            mutate(Season = "Spring"),
          read_csv(here("Data","wave_height_2023.csv")) %>%
            mutate(Date = make_datetime(YY, MM, DD, hh, mm)) %>%
            filter(Date >= as.Date("2023-10-21") & 
            Date <= as.Date("2023-11-03")) %>%
            mutate(Season = "Fall"))

stopifnot(load_ok)
```

# Cleaning & Pre-processing

We declare variable sets for identifiers, water chemistry, nutrients, and fDOM components. We retain only **Dawn** and **Dusk** samples, coerce numeric columns, set factor orders, drop NA grouping levels, and apply the **1.5×IQR rule** as a conservative outlier filter (replacing outliers with `NA`).

```{r clean-variables}

# Identifier variables
id_vars <- c("UniqueID", "Season", "Benthos", "Quad_ID", "Day_Night", "Sampling_Date", "Tide")

# Water chemistry variables (from pH probe file)
water_chem_vars <- c("Salinity", "pH", "TempInSitu", "TA")

# Nutrient variables (from pH probe file)
nutrient_vars <- c("NO3", "PO4", "SIL", "NO2", "NH4", "NOx")

# fDOM variables (from fDOM file)
#fdom_vars <- c("UVHumic", "Tyrosine", "VisibleHumic", "MarineHumic",
#               "Tryptophan", "FulvicAcid", "BIX", "HIX", "FI", "M_to_C")

# reduced variables
fdom_vars <- c("Humic","Prot", "BIX", "HIX", "M_to_C")

all_numeric_vars <- c(water_chem_vars, nutrient_vars, fdom_vars)

# only vars of interest

less_vars <- c("pH", "NOx", "NH4", "HIX")

# Helper to IQR-filter a numeric vector (returning NA for outliers)
iqr_filter <- function(x){
  if(!is.numeric(x)) return(x)
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lo <- q1 - 1.5 * iqr
  hi <- q3 + 1.5 * iqr
  ifelse(x < lo | x > hi, NA_real_, x)
}

# Cleaned dataset
data_clean <- data_merged %>%
  filter(Day_Night %in% keep_times) %>%
  select(-M_to_C) %>%
  mutate(M_to_C = MarineHumic / VisibleHumic) %>%
  select(all_of(id_vars), any_of(all_numeric_vars)) %>%
  mutate(across(all_of(all_numeric_vars), as.numeric)) %>%
  mutate(
    Benthos   = factor(Benthos, levels = c(control_label, "Rockweed", "Surfgrass", "Barnacle", "Mussel")),
    Day_Night = factor(Day_Night, levels = c("Dawn", "Dusk"))
  ) %>%
  drop_na(Benthos, Day_Night) %>%
  mutate(across(all_of(all_numeric_vars), iqr_filter)) %>% #iqr filter
  mutate(sample_type = if_else(Benthos == control_label,
                                   "Surface Water", "Foundation Species"))

levels(data_clean$Benthos) <- c("Surface Water", "Rockweed", "Surfgrass", "Barnacle", "Mussel") # change OO label
```

## Cleaning variables
Start by assessing assumptions of normality and homogeneity of variance.
```{r normality, fig.keep = "none"}
#-----short format df-----#
data_normal <- data_clean %>%
  mutate(HIX_transformed = log10(HIX),
         NOx_transformed = log10(NOx)) %>%
  select(-NOx, -HIX) %>%
  rename(NOx = NOx_transformed,
         HIX = HIX_transformed)

#-----long format df------#
testing_df <- data_clean %>%
  pivot_longer(cols = all_of(less_vars),
               names_to = "variable", values_to = "Value")

# running lm and creating diagnostic plots
map_dfr(less_vars, function(v){
  df_sub <- testing_df %>%
    filter(variable == v) # filter by variable
    options(contrasts = c("contr.sum", "contr.poly"))
    fit <- lm(Value ~ Day_Night * Benthos * Season, data = df_sub)
  par(mfrow = c(2,2)) # create 2x2 for plots
  plot(fit, main = paste("Diagnostics for", v)) # check assumptions using qq and residuals plots
})

# NOx and HIX looks like they need to be transformed
transform_var <- c("NOx", "HIX")
transformed_var <- map_dfr(transform_var, function(v){
  testing_df %>%
    filter(variable == v) %>% # filter by variable
    mutate(Value = sqrt(Value),
           variable = paste0(v,"_transformed"))
})

# check that transformations worked
transformed_vars <- c("NOx_transformed", "HIX_transformed")
map_dfr(transformed_vars, function(v){
  df_sub <- transformed_var %>%
    filter(variable == v) # filter by variable
    options(contrasts = c("contr.sum", "contr.poly"))
    fit <- lm(Value ~ Day_Night * Benthos * Season, data = df_sub)
  par(mfrow = c(2,2)) # create 2x2 for plots
  plot(fit, main = paste("Diagnostics for", v)) # check assumptions using qq and residuals plots
})

testing_df_norm <- bind_rows(testing_df, transformed_var) 
numeric_vars_normal <- c("pH", "NOx_transformed", "NH4", "HIX_transformed")
```

# Summary Statistics
Per habitat × diel period (and Season), we compute mean, SD, and sample size for all numeric variables.

```{r summary-stats}
data_summary <- data_clean %>%
  group_by(Benthos, Day_Night, Season) %>%
  summarise(
    across(all_of(all_numeric_vars), list(mean = ~ mean(., na.rm = TRUE),
                                          sd   = ~ sd(., na.rm = TRUE),
                                          n    = ~ sum(!is.na(.))
    )),
    .groups = "drop"
  )

data_summary %>%
  kable()
```

# Background plots

## PCA
Standardized within season. Do different seasons have distinct diel chemical/fDOM signatures? Which variables drive those differences?
```{r pca, fig.height=6}
pca_biplot <- function(df,
                       vars,
                       season_var = "Season",
                       color_var = "Day_Night",
                       shape_var = "Day_Night",
                       ellipse_var = "Benthos",
                       control_label = "Surface Water",
                       permanova_permutations = 999,
                       seed = 123) {

  # select vars of interest
  df_sub <- df %>%
    select(all_of(c(vars, season_var, color_var, ellipse_var))) %>%
    drop_na(all_of(vars))

  # standardize within season
  df_scaled <- df_sub %>%
    group_by(.data[[season_var]]) %>%
    mutate(across(all_of(vars), scale)) %>%
    ungroup()

  # run PCA
  num_mat <- df_scaled %>%
    select(all_of(vars)) %>%
    as.matrix()

  pca_res <- prcomp(num_mat, scale. = FALSE)

  # get scores
  scores <- as.data.frame(pca_res$x[, 1:2])
  colnames(scores) <- c("PC1", "PC2")

  scores <- scores %>%
    bind_cols(
      df_scaled %>%
        select(all_of(c(season_var, color_var, ellipse_var)))
    ) %>%
    mutate(Day_Night = factor(Day_Night, levels = c("Dawn", "Dusk")))

  # get loadings
  loadings <- as.data.frame(pca_res$rotation[, 1:2])
  colnames(loadings) <- c("PC1", "PC2")
  loadings$var <- rownames(loadings)

  # scale loadings
  mult <- min(
    diff(range(scores$PC1)) / diff(range(loadings$PC1)),
    diff(range(scores$PC2)) / diff(range(loadings$PC2))
  ) * 0.7

  loadings <- loadings %>%
    mutate(
      PC1 = PC1 * mult,
      PC2 = PC2 * mult
    )

  # variance
  var_expl <- (pca_res$sdev^2 / sum(pca_res$sdev^2)) * 100

  # medians for plot
  med <- scores %>%
    group_by_at(vars(one_of(c(season_var, color_var, shape_var)), "sample_type")) %>%
    summarise(
      PC1 = median(PC1, na.rm = TRUE),
      PC2 = median(PC2, na.rm = TRUE),
      .groups = "drop"
    )
    
  # permanova and dispersions
  pca_scores <- scores %>%
    filter(sample_type == "Foundation Species") %>%
    select(
      PC1,
      PC2,
      Day_Night = all_of(color_var),
      Season   = all_of(season_var)
    )

  set.seed(seed)
  
  permanova_res <- adonis2(
    pca_scores[, c("PC1","PC2")] ~ Day_Night * Season,
    data = pca_scores,
    method = "euclidean",
    permutations = permanova_permutations,
    by = "margin"
  )

  dist_mat <- dist(pca_scores[, c("PC1", "PC2")])

  disp_daynight <- betadisper(dist_mat, group = pca_scores$Day_Night)
  disp_season   <- betadisper(dist_mat, group = pca_scores$Season)

  dispersion_tests <- list(
    day_night = list(
      betadisper = disp_daynight,
      anova = anova(disp_daynight),
      permutest = permutest(disp_daynight, permutations = permanova_permutations)
    ),
    season = list(
      betadisper = disp_season,
      anova = anova(disp_season),
      permutest = permutest(disp_season, permutations = permanova_permutations)
    )
  )
  
  within_season <- pca_scores %>%
  group_split(Season) %>%
  lapply(function(d) {
    if (n_distinct(d$Day_Night) < 2) return(NULL)
    ad <- adonis2(
      d[, c("PC1","PC2")] ~ Day_Night,
      data = d,
      method = "euclidean",
      permutations = 999
    )
    data.frame(
      Season = unique(d$Season),
      R2 = ad$R2[1],
      F = ad$F[1],
      p = ad$`Pr(>F)`[1]
    )
  }) %>%
  bind_rows()

  # plot
    p <-
    ggplot() +

    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +

    # foundation species dots
    geom_point(
      data = scores %>% filter(sample_type == "Foundation Species"),
      aes_string(x = "PC1", y = "PC2", color = color_var, shape = shape_var),
      size = 3,
      alpha = 0.35
    ) +

    # median darker dot
    geom_point(
      data = med %>% filter(sample_type == "Foundation Species"),
      aes_string(x = "PC1", y = "PC2", color = color_var, shape = shape_var),
      size = 4.8,
      alpha = 1
    ) +

    # ellipses
    stat_ellipse(
      data = scores %>% filter(sample_type == "Foundation Species"),
      aes_string(x = "PC1", y = "PC2", group = color_var, color = color_var),
      linewidth = 1,
      alpha = 0.45,
      level = 0.95
    ) +

    # shape
    scale_shape_manual(values = c(16, 17), name = "Sampling Time") +

    # new shape for controls
    ggnewscale::new_scale("shape") +
    
    # control median
    geom_point(
      data = med %>% filter(sample_type == "Surface Water"),
      aes_string(x = "PC1", y = "PC2", color = color_var, shape = shape_var),
      size = 4.8,
      stroke = 1.5,
      alpha = 1,
      show.legend = FALSE
    ) +
    
    # control points
    geom_point(
      data = scores %>% filter(sample_type == "Surface Water"),
         aes_string(
             x = "PC1",
             y = "PC2",
            shape = shape_var,
            color = color_var
          ),
        fill  = NA,         
        size = 3,
        stroke = 1.5,         # outline thickness
        alpha = .35,
        show.legend = FALSE
    ) +
      
    # open shapes for controls
    scale_shape_manual(values = c(21, 24)) +
      
    # colors for all points
    scale_color_manual(
      values = c("Dawn" = "#FDB813", "Dusk" = "#003865"),
      name = "Sampling Time"
    ) +

    # loadings
    geom_segment(
      data = loadings,
      aes(x = 0, y = 0, xend = PC1, yend = PC2),
      arrow = arrow(length = unit(0.2, "cm")),
      inherit.aes = FALSE
    ) +
      
    geom_text_repel(
      data = loadings,
      aes(x = PC1, y = PC2, label = var),
      size = 4,
      point.padding = 1,
      box.padding = 0.75,
      segment.color = "grey50",
      inherit.aes = FALSE
    ) +

    facet_wrap(as.formula(paste("~", season_var))) +

    labs(
      x = paste0("PC1 (", round(var_expl[1], 1), "%)"),
      y = paste0("PC2 (", round(var_expl[2], 1), "%)")
    ) +
    
    theme_bw()+
    theme(strip.background = element_rect(fill = "white"))
    
  return(list(
    plot = p,
    permanova = permanova_res,
    within_season = within_season,
    dispersion_tests = dispersion_tests
  ))
}

pca_biplot(data_normal,
  vars      = less_vars,
  season_var = "Season",
  color_var = "Day_Night",
  shape_var = "Day_Night",
  ellipse_var = "Benthos")

# run PCA on transformed variables ~
```
## Map
```{r}
states<-map_data("state") #all state coordinates
ca_data <- states %>%
  filter(region == "california") #just filter out california

ggplot()+
  geom_polygon(data = ca_data,
               aes(x=long, y=lat, group = group), color = "navyblue",
               fill = "white", linewidth = 1)+
  geom_point(aes(x = -118.294363, y = 33.704647), color= "navyblue", size = 5)+
  coord_map()+
  theme_void()
```

## Temperature plots
### Using HOBO data
Lower mean temperatures in Spring, but higher extreme temperatures/max.
```{r temp, fig.width = 7, fig.height = 5}
# overall temperatures
ggplot(temperature, 
       aes(x = datetime, y = temperature_c, color = Benthos)) +
  geom_line(alpha = 0.6, linewidth = 0.4) +
  labs(x = "Date", y = "Temperature (°C)") +
  scale_color_manual(values = c(
                     "Surfgrass" = "#5BB360",
                     "Rockweed" = "#5A8F5D",
                     "Barnacle" = "#6E86C2",
                     "Mussel" = "#233B75")) +
  facet_wrap(Season~Benthos, 
             scale = "free_x",
             ncol = 4)+
  theme_minimal()

# not faceted
ggplot(temperature, aes(x = datetime, y = temperature_c, color = Season)) +
  geom_line(alpha = 0.6, linewidth = 0.4) +
  labs(x = NULL, y = "Temperature (°C)") +
  scale_color_manual(values = c(
    "Fall"   = "#D95F02",
    "Spring" = "#1B9E77"
  )) +
  facet_wrap(~Season, 
             scale = "free_x",
             ncol = 1)+
  theme_classic()+
  theme(legend.position = "none",
        strip.background = element_blank(),
        text = element_text(size = 20),
        plot.margin = unit(c(0,1,0.5,0.5), "cm"))

# daily temperatures
temperature_summary <- temperature %>%
  mutate(Day = as.Date(datetime)) %>%
  group_by(Season, Benthos, Day) %>%
  summarise(
    temp_mean = mean(temperature_c, na.rm = TRUE),
    temp_sd   = sd(temperature_c, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(temperature_summary, 
       aes(Day, temp_mean, color = Benthos, fill = Benthos)) +
  geom_ribbon(aes(ymin = temp_mean - temp_sd,
                  ymax = temp_mean + temp_sd),
              alpha = 0.2, linewidth = 0) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c(
                     "Surfgrass" = "#5BB360",
                     "Rockweed" = "#5A8F5D",
                     "Barnacle" = "#6E86C2",
                     "Mussel" = "#233B75")) +
  scale_fill_manual(values = c(
                     "Surfgrass" = "#5BB360",
                     "Rockweed" = "#5A8F5D",
                     "Barnacle" = "#6E86C2",
                     "Mussel" = "#233B75")) +
  facet_wrap(Season~Benthos, 
             scales = "free_x",
             ncol = 4) +
  labs(x = NULL, y = "Temperature (°C)") +
  theme_minimal()

# statistical difference between seasons, but not benthos
tempmodel <- aov(temp_mean ~ Season * Benthos, 
                 data = temperature_summary)
summary(tempmodel)

# violin plot
ggplot(temperature,
       aes(x = Season, y = temperature_c, fill = Season)) +
  geom_violin(trim = FALSE, adjust = 1.2, alpha = 0.6) +
  scale_fill_manual(values = c(
    "Fall"   = "#D95F02",
    "Spring" = "#1B9E77"
  )) +
  facet_wrap(~ Benthos) +
  labs(
    y = "Temperature (°C)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )
```

### Using buoy data
```{r}
tempplot <- ggplot(temp,
       aes(x = Season, y = WTMP, fill = Season)) +
  geom_violin(trim = FALSE, adjust = 1.2, alpha = 0.6) +
  scale_fill_manual(values = c(
    "Fall"   = "#D95F02",
    "Spring" = "#1B9E77"
  )) +
  labs(
    y = "Temperature (°C)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    text = element_text(size = 20),
    strip.text = element_text(face = "bold")
  )
```


## CUTI plot
```{r cuti, fig.width = 8, fig.height = 5}
ggplot(cuti, aes(x = Date, y = index, color = Season)) +
  geom_line(alpha = 1, linewidth = 1) +
  geom_point()+
  labs(x = NULL, y = "CUTI") +
  scale_color_manual(values = c(
    "Fall"   = "#D95F02",
    "Spring" = "#1B9E77"
  )) +
  facet_wrap(~Season, 
             scale = "free_x",
             ncol = 1)+
  theme_classic()+
  theme(legend.position = "none",
        strip.background = element_blank(),
        text = element_text(size = 20),
        plot.margin = unit(c(0,1,0.5,0.5), "cm"))

cutiplot <- ggplot(cuti, aes(x = Season, y = index, fill = Season)) +
  geom_violin(trim = FALSE, adjust = 1.2, alpha = 0.6) +
  scale_fill_manual(values = c(
    "Fall"   = "#D95F02",
    "Spring" = "#1B9E77"
  )) +
  labs(
    y = "CUTI"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    text = element_text(size = 20),
    strip.text = element_text(face = "bold"))

tempplot + cutiplot + plot_layout(axes = "collect")

```

## Light plots
More light in spring, and barnacle and mussels higher on average than rockweeds and surfgrass.
```{r light, fig.width = 10}
# overall light lux
ggplot(temperature, 
       aes(x = datetime, y = light_lux, color = Benthos)) +
  geom_line(alpha = 0.6, linewidth = 0.4) +
  labs(y = "Light (LUX)") +
  scale_color_manual(values = c(
                     "Surfgrass" = "#5BB360",
                     "Rockweed" = "#5A8F5D",
                     "Barnacle" = "#6E86C2",
                     "Mussel" = "#233B75")) +
  facet_wrap(Season~Benthos, 
             scale = "free_x",
             ncol = 4)+
  theme_minimal()

# daily light lux
light_summary <- temperature %>%
  mutate(Day = as.Date(datetime)) %>%
  group_by(Season, Benthos, Day) %>%
  summarise(
    light_mean = mean(light_lux, na.rm = TRUE),
    light_sd   = sd(light_lux, na.rm = TRUE),
    .groups = "drop"
  )

# plot
ggplot(light_summary, 
       aes(Day, light_mean, color = Benthos, fill = Benthos)) +
  geom_ribbon(aes(ymin = light_mean - light_sd,
                  ymax = light_mean + light_sd),
              alpha = 0.2, linewidth = 0) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c(
                     "Surfgrass" = "#5BB360",
                     "Rockweed" = "#5A8F5D",
                     "Barnacle" = "#6E86C2",
                     "Mussel" = "#233B75")) +
  scale_fill_manual(values = c(
                     "Surfgrass" = "#5BB360",
                     "Rockweed" = "#5A8F5D",
                     "Barnacle" = "#6E86C2",
                     "Mussel" = "#233B75")) +
  facet_wrap(Season~Benthos, 
             scales = "free_x",
             ncol = 4) +
  labs(x = NULL, y = "Light (LUX)") +
  theme_minimal()

# difference between season and benthos
lightmodel <- aov(light_mean ~ Season * Benthos, 
                 data = light_summary)
summary(lightmodel)

# violin plot
ggplot(temperature,
       aes(x = Season, y = light_lux, fill = Season)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  scale_fill_manual(values = c(
    "Fall"   = "#D95F02",
    "Spring" = "#1B9E77"
  )) +
  facet_wrap(~ Benthos, scales = "free_y") +
  scale_y_log10() +
  labs(
    y = "Light (lux, log scale)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )

```

# pH and Nitrogen - "Upwelling drives primary production"
Note: removed two outliers
```{r ph-nit}
# pH and nitrate relationship
ggplot(data_clean %>%
         filter(NOx > 0.07, pH >7.71), aes(x = NOx, y = pH)) +
  coord_trans(x = "log10") +
  geom_point(alpha = 0.5, aes(color = Season, shape = Season)) +
  scale_color_manual(values = c(
    "Fall"   = "#D95F02",
    "Spring" = "#1B9E77"
  )) +
  facet_wrap(~Day_Night, scale = "free")+
  geom_smooth(method = "lm",
              se = TRUE, aes(color = "black")) +
  theme(strip.background = element_rect(fill = "white"))

## with benthos
ggplot(data_clean %>%
         filter(NOx > 0.07, pH >7.71), aes(x = NOx, y = pH)) +
  coord_trans(x = "log10") +
  geom_point(alpha = 0.5, aes(color = Season, shape = Season)) +
  scale_color_manual(values = c(
    "Fall"   = "#D95F02",
    "Spring" = "#1B9E77"
  )) +
  geom_smooth(method = "lm", se = TRUE, aes(color = "black")) +
  facet_wrap(Benthos~Day_Night, scale = "free", ncol = 5)+
  theme(strip.background = element_rect(fill = "white"))

```

### Model
```{r ph-nit-model, fig.height=5}
dat2 <- data_clean %>%
  mutate(
    Day_Night = factor(Day_Night),
    Season    = factor(Season),
    Benthos   = factor(Benthos),
    logNOx    = log10(NOx)
  ) %>%
  filter(is.finite(NOx), NOx > 0, NOx > 0.07, pH >7.71) # remove outliers

## We expect the relationship between pH and NOx to change with Day_Night. pH changes with season and benthos

pHNOxmodel <- lmer(pH ~ logNOx * Day_Night * Season + (1|Benthos), data = dat2)
Anova(pHNOxmodel, type = "III")
summary(pHNOxmodel)

emm_slopes <- emmeans(pHNOxmodel, ~ Day_Night, var = "logNOx", 
                      infer = c(TRUE, TRUE)) 
emm_slopes # dawn significant, dusk is not

contrast(emm_slopes, method = "pairwise")
```

### Density plot
```{r ph-nit-plot, fig.height=6, fig.width=12}
ggplot(data_clean %>% filter(NOx > 0.07, pH >7.71), # remove outliers
       aes(x = NOx, y = pH)) +
  # create open circles for ctrl
  geom_point(data = data_clean %>% 
               filter(sample_type == "Surface Water"),
            aes(color = Season, shape = Season), stroke = 1.2, alpha = 0.4) + scale_shape_manual(values = c(21, 24)) +
  # new shape scale
  ggnewscale::new_scale("shape") +
  # filled in circles for benthos
  geom_point(data = data_clean %>% 
               filter(sample_type == "Foundation Species"),
            aes(color = Season, shape = Season), alpha = 0.4) +
  scale_shape_manual(values = c(16, 17)) +
  # color by season
  scale_color_manual(values = c("Fall" = "#D95F02", 
                                "Spring" = "#1B9E77")) +
  # density contours filled by season
  stat_density_2d(aes(fill = Season, group = Season),
    geom = "polygon",
    bins = 6,
    n = 200,
    alpha = 0.3) +
    scale_fill_manual(values = c("Fall" = "#D95F02", 
                                "Spring" = "#1B9E77")) +
  # log transform x axis
  scale_x_log10(limits = c(0.07, 10))+
  facet_wrap(~Day_Night)+
  labs(y = expression(paste("pH"[T])),
       x = expression(paste("log"[10],"(NO"[x],")")))+
  theme(strip.background = element_rect(fill = "white"),
        text = element_text(size = 20))

```

```{r hix-bix, echo = FALSE, eval = FALSE}
# HIX and BIX are correlated with each other, makes sense. There is either more fresh/labile DOM (BIX) or more complex DOM (HIX). 
ggplot(data_clean %>%
         mutate(Benthos = factor(Benthos, levels = c("Surface Water", "Barnacle", "Mussel", "Rockweed", "Surfgrass"))),
       aes(x = BIX, y = HIX, color = Day_Night)) +
  geom_point() +
  scale_color_manual(values = c("Dawn" = "#FDB813", "Dusk" = "#003865")) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(
    rows = vars(Season),
    cols = vars(Benthos)
  ) +
  labs(
    color = "Sampling Time"
  )

# relationships between nitrates and BIX
## consumer dawn: as NOx decreases, BIX increases
ggplot(data_clean  %>%
         mutate(Benthos = factor(Benthos, levels = c("Surface Water", "Barnacle", "Mussel", "Rockweed", "Surfgrass"))),
       aes(x = NOx, y = BIX, color = Day_Night)) +
  geom_point() +
  scale_color_manual(values = c("Dawn" = "#FDB813", "Dusk" = "#003865")) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid2(
    rows = vars(Season),
    cols = vars(Benthos),
    scales = "free_x"
  ) +
  labs(
    color = "Sampling Time"
  )

## consumer dawn: as NOx increases, HIX increases
ggplot(data_clean  %>%
         mutate(Benthos = factor(Benthos, levels = c("Surface Water", "Barnacle", "Mussel", "Rockweed", "Surfgrass"))),
       aes(x = NOx, y = HIX, color = Day_Night)) +
  geom_point() +
  scale_color_manual(values = c("Dawn" = "#FDB813", "Dusk" = "#003865")) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(
    rows = vars(Season),
    cols = vars(Benthos),
    scales = "free_x"
  ) +
  labs(
    color = "Sampling Time"
  )

# relationships between ammonium and HIX/BIX
## consumer dawn: as NH4 increases, HIX increases
#nh4hixplot<-
ggplot(data_clean, aes(x = NH4, y = HIX, color = Day_Night)) +
  geom_point() +
  scale_color_manual(values = c("Dawn" = "#FDB813", "Dusk" = "#003865")) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid2(
    rows = vars(Season),
    cols = vars(Benthos), 
    scales = "free",
    switch = "y")


## consumer dawn: as NH4 increases, BIX decreases; rockweed increases at dawn
nh4bixplot <-
ggplot(data_clean, aes(x = NH4, y = BIX, color = Day_Night)) +
  geom_point() +
  scale_color_manual(values = c("Dawn" = "#FDB813", "Dusk" = "#003865")) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid2(
    rows = vars(Season),
    cols = vars(Benthos), 
    scales = "free",
    switch = "y")
  
nh4hixplot + nh4bixplot + plot_layout(guides = "collect", ncol = 1, 
                                      axes = "collect")
# combine benthos, facet by sampling time
dielhixplot <-
ggplot(data_clean  %>% filter(Season == "Spring") %>%
         mutate(Benthos = factor(Benthos, levels = c("Surface Water", "Barnacle", "Mussel", "Rockweed", "Surfgrass"))),
       aes(x = NOx, y = HIX, color = Benthos)) +
  geom_point() +
  scale_color_manual(values = c("Surface Water" = "#8C8C8C", 
                     "Surfgrass" = "#5BB360",
                     "Rockweed" = "#5A8F5D",
                     "Barnacle" = "#6E86C2",
                     "Mussel" = "#233B75")) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~Day_Night)+
  labs(
    color = "Benthos"
  )

dielbixplot <-
ggplot(data_clean  %>% filter(Season == "Spring") %>%
         mutate(Benthos = factor(Benthos, levels = c("Surface Water", "Barnacle", "Mussel", "Rockweed", "Surfgrass"))),
       aes(x = NOx, y = BIX, color = Benthos)) +
  geom_point() +
  scale_color_manual(values = c("Surface Water" = "#8C8C8C", 
                     "Surfgrass" = "#5BB360",
                     "Rockweed" = "#5A8F5D",
                     "Barnacle" = "#6E86C2",
                     "Mussel" = "#233B75")) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~Day_Night)+
  labs(
    color = "Benthos"
  )

dielhixplot + dielbixplot + plot_layout(guides = "collect")

ggsave(here("Output","HIXBIXPlot.png"), width = 12, height = 6)
```

```{r nh4-nox, eval = FALSE}
# separated by benthos
ggplot(data_clean %>% 
         mutate(Benthos = factor(Benthos, levels = c("Surface Water", "Barnacle", "Mussel", "Rockweed", "Surfgrass"))), 
       aes(x = NH4, y = NOx)) +
  coord_trans(x = "log10", y = "log10")+
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +   
  #scale_color_manual(values = c("Dawn" = "#FDB813", "Dusk" = "#003865")) + 
  facet_grid2(Season~Benthos, scales = "free", switch = "y", independent = "all" ) + 
  labs( x = expression(paste("NH"[4]," (", mu, "mol L"^-1,")")), 
        y = expression(paste("NO"[x]," (", mu, "mol L"^-1,")")), 
        color = "Sampling Time" )

# combined
ggplot(data_bench %>% 
         mutate(Benthos = factor(Benthos, levels = c("Surface Water", "Barnacle", "Mussel", "Rockweed", "Surfgrass"))), 
       aes(x = NH4, y = NOx, color = Sample_type)) +
  coord_trans(x = "log10", y = "log10")+
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +   
  #scale_color_manual(values = c("Dawn" = "#FDB813", "Dusk" = "#003865")) + 
  facet_grid2(Season~., switch = "y") + 
  labs( x = expression(paste("NH"[4]," (", mu, "mol L"^-1,")")), 
        y = expression(paste("NO"[x]," (", mu, "mol L"^-1,")")), 
        color = "Sampling Time" )

# separate OO from foundation species
data_bench <- data_logged %>%
  mutate(Sample_type = if_else(Benthos == "Surface Water", "Surface Water", "Foundation Species"))

# we tested random effects, daynight and benthos = 0
nitmodelfall <- lm(logNOx ~ logNH4 * Sample_type, 
                 data = data_bench %>%
                   filter(Season == "Fall"))
anova(nitmodelfall)
ranef(nitmodelfall)
summary(nitmodelfall)


nitmodelspring <- lmer(logNOx ~ logNH4 * Sample_type + (1|Day_Night), 
                 data = data_bench %>%
                   filter(Season == "Spring"))
anova(nitmodelspring)
ranef(nitmodelspring)
```

```{r nh4-nox-sig, eval = FALSE}
# ONLY KEEPING SIG LINES

## create log columns
data_logged <- data_clean %>%
  mutate(
    logNH4 = log10(NH4),
    logNOx = log10(NOx),
    Season = as.factor(Season)
  )

## nitrogen model
nitrogenmodel <- lm(logNOx ~ logNH4 * Season * Benthos,
  data = data_logged
)
Anova(nitrogenmodel, type = "III")

## calculate slopes for day and night separately within each benthos and season
emm <- summary(emtrends(nitrogenmodel,
  ~ Benthos * Season,
  var = "logNH4"),
  infer = c(TRUE,TRUE)) %>%
  as.data.frame()

print(emm)

## filter sig slopes
alpha <- 0.05

sig_slopes <- emm %>%
  filter(p.value < alpha) %>%
  transmute(
    Day_Night, Benthos, Season,
    slope = `logNH4.trend`
  )

## intercept on log scale for each panel at log(NH4)=0
new0 <- data_logged %>%
  distinct(Day_Night, Benthos, Season) %>%
  mutate(logNH4 = 0)

new0$intercept <- predict(nitrogenmodel, newdata = new0)

line_df <- new0 %>%
  left_join(sig_slopes, by = c("Day_Night", "Benthos", "Season")) %>%
  filter(!is.na(slope))

## plot
ggplot(data_logged %>%
    mutate(Benthos = factor(Benthos, levels = c("Surface Water", "Barnacle", "Mussel", "Rockweed", "Surfgrass"))),
  aes(x = logNH4, y = logNOx, color = Day_Night)
) +
  geom_point() +
  geom_abline(
    data = line_df,
    aes(intercept = intercept, slope = slope, color = Day_Night),
    inherit.aes = FALSE,
    linewidth = 1
  ) +
  scale_color_manual(values = c("Dawn" = "#FDB813", "Dusk" = "#003865")) +
  facet_grid2(Season ~ Benthos, 
              scales = "free", 
              switch = "y") +
  labs(
    x = expression(paste("log(NH"[4], ") (", mu, "mol L"^-1, ")")),
    y = expression(paste("log(NO"[x], ") (", mu, "mol L"^-1, ")")),
    color = "Sampling Time"
  )
```

```{r nh4-nox-sep-sig, eval = FALSE}
## create log columns
data_logged <- data_clean %>%
  mutate(
    logNH4 = log10(NH4),
    logNOx = log10(NOx),
    Season= as.factor(Season)
  )

## fit one model per benthos x season and store
## subset by benthos but keep Surface Water to compare emmeans
### random effect of dawn/dusk
models <- set_names(levels(data_logged$Benthos)) %>%
  map(function(b) {
    dat_b <- filter(data_logged, Benthos == b)
    tryCatch(
      lmer(logNOx ~ logNH4 * Benthos + (1|Day_Night), data = dat_b),
      error = function(e) {
        warning(glue::glue("Model failed for Benthos = '{b}': {e$message}"))
        NULL
      }
    )
  })

## extract emtrends for each model: Day_Night | Season 
emm_all <- imap_dfr(models, function(mod, b) {
  em <- tryCatch(
    summary(emtrends(mod, ~ Season, var = "logNH4"), infer = c(TRUE, TRUE)) %>%
      as.data.frame(),
    error = function(e) {
      warning(glue::glue("emtrends failed for '{b}': {e$message}"))
      return(NULL)
    }
  )
  if (is.null(em)) return(tibble())
  em %>% mutate(Benthos = b) %>% select(Benthos, everything())
})

emm_all %>% 
  kable()

## filter significant slopes
alpha <- 0.05
sig_slopes <- emm_all %>%
  filter(p.value <= alpha) %>%
  transmute(
    Benthos,
    Season,
    Day_Night,
    slope = `logNH4.trend`,
    slope_se = SE,
    slope_p = p.value
  )

sig_slopes %>%
  kable()

## build intercepts for each distinct panel (logNH4 = 0)
new0 <- data_logged %>%
  distinct(Benthos, Day_Night, Season) %>%
  mutate(logNH4 = 0)

## predict intercepts using the correct Benthos model
# pmap over the rows to call predict with the right model
intercepts <- new0 %>%
  rowwise() %>%
  mutate(
    intercept = {
      mod <- models[[as.character(Benthos)]]
      if (is.null(mod)) NA_real_ else {
        # pull the data actually used to fit this model
        mf <- model.frame(mod)

        newd <- tibble(
          logNH4 = 0,
          Day_Night = factor(as.character(Day_Night), levels = levels(mf$Day_Night)),
          Season    = factor(as.character(Season),    levels = levels(mf$Season))
        )

        as.numeric(predict(mod, newdata = newd))
      }
    }
  ) %>%
  ungroup()

## join slopes -> only keep rows where slope is significant (non-NA)
line_df <- intercepts %>%
  left_join(sig_slopes, by = c("Benthos", "Season", "Day_Night")) %>%
  filter(!is.na(slope))

## plot: points + only-significant lines

ggplot(data_logged,
       aes(x = logNH4, y = logNOx, color = Day_Night)) +
  geom_point() +
  geom_abline(
    data = line_df,
    aes(intercept = intercept, slope = slope, color = Day_Night),
    inherit.aes = FALSE,
    linewidth = 1
  ) +
  scale_color_manual(values = c("Dawn" = "#FDB813", "Dusk" = "#003865")) +
  facet_grid2(Season ~ Benthos, scales = "free", switch = "y") +
  labs(
    x = expression(paste("log(NH"[4], ") (", mu, "mol L"^-1, ")")),
    y = expression(paste("log(NO"[x], ") (", mu, "mol L"^-1, ")")),
    color = "Sampling Time"
  ) +
  theme_minimal()

```

```{r pH-HIX, eval=FALSE}
## new plot with foundation species vs OO
ggplot(data_bench %>% 
         mutate(Benthos = factor(Benthos, levels = c("Surface Water", "Barnacle", "Mussel", "Rockweed", "Surfgrass"))), 
       aes(x = pH, y = HIX, color = Sample_type)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +   
  #scale_color_manual(values = c("Dawn" = "#FDB813", "Dusk" = "#003865")) + 
  facet_grid2(Season~., switch = "y") + 
  labs( x = expression(paste("pH")), 
        y = expression(paste("HIX")), 
        color = "Sampling Time" )

# pH and HIX plot
ggplot(data_clean, 
       aes(x = pH, y = HIX, color = Benthos)) +
  geom_point() +
  scale_color_manual(values = c("Surface Water" = "#8C8C8C", 
                     "Surfgrass" = "#5BB360",
                     "Rockweed" = "#5A8F5D",
                     "Barnacle" = "#6E86C2",
                     "Mussel" = "#233B75")) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = expression(paste("pH")),
    y = expression(paste("HIX")),
    color = "Benthos"
  )+ 
  facet_grid2(
    rows = vars(Season),
    cols = vars(Benthos),
    switch = "y"
    #scale = "free",
    #independent = "all"
  )

#ggsave(here("Output","pH_HIX_plot.png"), width = 10, height = 8)


## updated model for species vs OO
HIXmodelfall <- lmer(HIX ~ pH * Sample_type + (1|Benthos), data = data_bench %>%
                       filter(Season == "Fall"))
anova(HIXmodelfall)
ranef(HIXmodelfall)

HIXmodelspring <- lmer(HIX ~ pH * Sample_type + (1|Benthos), data = data_bench %>%
                       filter(Season == "Spring"))

anova(HIXmodelspring)
ranef(HIXmodelspring)
```

```{r pH-HIX-plot, eval=FALSE}
# start by fitting the model to see significant trends
pHHIXmodel <- lmer(HIX ~ pH + (1|Day_Night), data = data_clean)
Anova(pHHIXmodel, type = "III")

pHHIXmodelrw <- lmer(HIX ~ pH + (1|Day_Night), data = data_clean %>%
                     filter(Benthos == "Mussel", Season == "Fall"))

anova (pHHIXmodelrw)

# only filter out significant slopes
alpha <- 0.05

tr_SB <- summary(emtrends(pHHIXmodel, 
                          specs = ~ Season * Benthos, 
                          var = "pH"),
                infer = c(TRUE, TRUE)) %>% # get p values
        as.data.frame()

print(tr_SB)

sig_tr_SB <- tr_SB %>%
  filter(p.value < alpha) %>%
  transmute(
    Season,
    Benthos,
    slope = pH.trend
  )


# get intercepts
intercepts <- data_clean %>%
  distinct(Season, Benthos) %>%
  mutate(pH = 0)

intercepts$intercept <- predict(pHHIXmodel, newdata = intercepts)

line_df <- intercepts %>%
  left_join(sig_tr_SB, by = c("Season", "Benthos")) %>%
  filter(!is.na(slope))   # ONLY significant regressions

# plot sig lines
ggplot(
  data_clean,
  aes(x = pH, y = HIX, color = Benthos)
) +
  geom_point() +
  scale_color_manual(values = c(
    "Surface Water" = "#8C8C8C",
    "Surfgrass"  = "#5BB360",
    "Rockweed"   = "#5A8F5D",
    "Barnacle"   = "#6E86C2",
    "Mussel"     = "#233B75"
  )) +
  geom_abline(
    data = line_df,
    aes(intercept = intercept, slope = slope,
        color = Benthos),
    linewidth = 1
  ) +
  labs(
    x = expression(paste("pH")),
    y = expression(paste("HIX")),
    color = "Benthos"
  ) +
  facet_grid2(
    rows = vars(Season),
    cols = vars(Benthos),
    switch = "y"
  )

```

```{r pH-hix-combined, eval=FALSE}
ggplot(data_clean, 
       aes(x = pH, y = HIX)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = expression(paste("pH")),
    y = expression(paste("HIX")),
    color = "Benthos"
  )+ 
  facet_wrap(Benthos~Season, scale = "free", nrow = 2)

# lines colored by benthos
ggplot(data_clean, 
       aes(x = pH, y = HIX, color = Benthos)) +
  geom_point() +
  scale_color_manual(values = c("Surface Water" = "#8C8C8C", 
                     "Surfgrass" = "#5BB360",
                     "Rockweed" = "#5A8F5D",
                     "Barnacle" = "#6E86C2",
                     "Mussel" = "#233B75")) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = expression(paste("pH")),
    y = expression(paste("HIX")),
    color = "Benthos"
  )+ 
  facet_wrap (~Season)

# one trend line
ggplot(data_clean, 
       aes(x = pH, y = HIX)) +
  geom_point(aes(color = Benthos)) +
  scale_color_manual(values = c("Surface Water" = "#8C8C8C", 
                     "Surfgrass" = "#5BB360",
                     "Rockweed" = "#5A8F5D",
                     "Barnacle" = "#6E86C2",
                     "Mussel" = "#233B75")) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    x = expression(paste("pH")),
    y = expression(paste("HIX")),
    color = "Benthos"
  )+ 
  facet_wrap (~Season)
```

```{r pH-TA, eval=FALSE}
# original plot
ggplot(data_clean %>%
        mutate(Day_Night = factor(Day_Night, levels = c("Dusk", "Dawn"))), 
       aes(x = pH, y = TA, color = Benthos)) +
  geom_point() +
  scale_color_manual(values = c("Surface Water" = "#8C8C8C", 
                     "Surfgrass" = "#5BB360",
                     "Rockweed" = "#5A8F5D",
                     "Barnacle" = "#6E86C2",
                     "Mussel" = "#233B75")) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = expression(paste("pH")),
    y = expression(paste("TA")),
    color = "Benthos"
  )+ 
  facet_grid2(
    rows = vars(Season),
    cols = vars(Benthos),
    switch = "y"
    #scale = "free",
    #independent = "all"
  )
```

```{r pH-TA-plot, eval=FALSE}
# model
TAmodel <- lm(TA ~ pH * Benthos * Season, data = data_clean)
anova(TAmodel)

# plot significant lines
alpha <- 0.05

TA_slopes <- summary(emtrends(TAmodel, 
                          specs = ~ Season * Benthos, 
                          var = "pH"),
                infer = c(TRUE, TRUE)) %>% # get p values
                as.data.frame()

print(TA_slopes)

sig_TA_slopes <- TA_slopes %>%
  filter(p.value < alpha) %>%
  transmute(
    Season,
    Benthos,
    slope = pH.trend
  )

# get intercepts
intercepts <- data_clean %>%
  distinct(Season, Benthos) %>%
  mutate(pH = 0)

intercepts$intercept <- predict(TAmodel, newdata = intercepts)

line_df <- intercepts %>%
  left_join(sig_TA_slopes, by = c("Season", "Benthos")) %>%
  filter(!is.na(slope))   # ONLY significant regressions

# plot sig lines
ggplot(
  data_clean %>%
    mutate(Day_Night = factor(Day_Night, levels = c("Dusk", "Dawn"))),
  aes(x = pH, y = TA, color = Benthos)
) +
  geom_point() +
  scale_color_manual(values = c(
    "Surface Water" = "#8C8C8C",
    "Surfgrass"  = "#5BB360",
    "Rockweed"   = "#5A8F5D",
    "Barnacle"   = "#6E86C2",
    "Mussel"     = "#233B75"
  )) +
  geom_abline(
    data = line_df,
    aes(intercept = intercept, slope = slope,
        color = Benthos),
    linewidth = 1
  ) +
  labs(
    x = expression(paste("pH")),
    y = expression(paste("TA")),
    color = "Benthos"
  ) +
  facet_grid2(
    rows = vars(Season),
    cols = vars(Benthos),
    switch = "y"
  )
```

# Dawn vs dusk
## ANOVA results
```{r ANOVA-table}
# individual models depending on whether they have random effect

## pH model without random effect
ph_model <- lm(Value ~ Day_Night * Benthos * Season, 
            data = testing_df_norm %>% 
              filter(variable == "pH"))
Anova(ph_model, type = "III")

## NOx model without random effect
NOx_model <- lm(Value ~ Day_Night * Benthos * Season, 
            data = testing_df_norm %>% 
              filter(variable == "NOx"))
Anova(NOx_model, type = "III")

## NH4 model with random effect
NH4_model <- lmer(Value ~ Day_Night * Benthos * Season + (1|Quad_ID), 
            data = testing_df_norm %>% 
              filter(variable == "NH4"))
Anova(NH4_model)
summary(NH4_model)

## HIX model with random effect
HIX_model <- lmer(Value ~ Day_Night * Benthos * Season + (1|Quad_ID), 
            data = testing_df_norm %>% 
              filter(variable == "HIX"))
Anova(HIX_model)
summary(HIX_model)

```

## Diel variability summary statistics
```{r diel-percent-change}
## percent change per benthos
pct_change_df <- testing_df %>%
  filter(variable == "NOx") %>%
  group_by(Season, Benthos, Day_Night) %>%
  summarize(
    mean_val = mean(Value, na.rm = TRUE),
    n = dplyr::n(),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from  = Day_Night,
    values_from = c(mean_val, n),
    names_sep = "_"
  ) %>%
  mutate(
    diel_diff = mean_val_Dusk - mean_val_Dawn,
    pct_change = ifelse(mean_val_Dawn == 0, NA_real_,
                        100 * diel_diff / mean_val_Dawn),
    n_total = n_Dawn + n_Dusk
  )

# not including dawn vs dusk
seasonal_diff_df <- testing_df %>%
  filter(variable == "NOx") %>%
  group_by(Season, Benthos) %>%
  summarize(
    mean_val = mean(Value, na.rm = TRUE),
    n = dplyr::n(),
    .groups = "drop"
  )

```

Let's look at differences across benthos in their dawn/dusk values over the two seasons

## Diel boxplots with significance levels
```{r diel-boxplots}
build_sig_df <- function(var_name = less_vars, data = data_clean) {
  stopifnot(var_name %in% names(data))
  
  # fit ANOVA
  fml <- as.formula(paste(var_name, "~ Day_Night * Season * Benthos + (1|Quad_ID)"))
  fit <- lmer(fml, data = data)
  
  # Dawn vs Dusk within each Season × Benthos
  emm  <- emmeans(fit, ~ Day_Night | Season * Benthos)
  contr_df <- as.data.frame(contrast(emm, method = "revpairwise"))
  # columns: Season, Benthos, contrast, estimate, SE, df, t.ratio, p.value
  
  # Turn p-values into stars
  sig_df <- contr_df %>%
    select(Season, Benthos, p.value) %>%
    distinct() %>%
    mutate(sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ "ns"
    )) %>%
    left_join(
      data %>%
        group_by(Season, Benthos) %>%
        summarise(y_pos = max(.data[[var_name]], na.rm = TRUE),
                  .groups = "drop"),
      by = c("Season", "Benthos")
    ) %>%
    mutate(y_pos = y_pos * 1.05) %>% # a bit above top whisker
    filter(sig != "ns")
}

plot_diel_with_sig <- function(var_name, var_label, 
                               data = data_clean,
                               free_y = FALSE) {
  stopifnot(var_name %in% names(data))
  
  sig_df <- build_sig_df(var_name, data = data)
  
  facet_scales <- if (free_y) "free_y" else "fixed"
  
  plot <- ggplot(
    data %>% filter(!is.na(.data[[var_name]])),
    aes(x = Day_Night, y = .data[[var_name]], fill = Day_Night)
  ) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
    facet_grid(rows = vars(Season),cols = vars(Benthos),
              scales = facet_scales,switch = "y"
) +
    scale_fill_manual(values = c("Dawn" = "#FDB813", "Dusk" = "#003865")) +
    labs(
      #title = paste("Diel Patterns in", var_label),
      x = "Sampling Time",
      y = var_label
    ) +
    theme(
      legend.position = "none",
      strip.background = element_rect(fill = "grey90"),
      strip.placement = "outside",
      axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
    geom_text(
      data = sig_df,
      aes(x = 1.5, y = y_pos, label = sig),
      inherit.aes = FALSE,
      size = 4
    )
}

```

## Example plots

```{r diel-boxplot-ex}
(p_ph   <- plot_diel_with_sig("pH",   "pH (Total Scale)"))
(p_TA   <- plot_diel_with_sig("TA",   expression(paste("Total Alkalinity (µmol kg"^-1,")")))
  + theme(strip.text.y = element_blank(), strip.background.y = element_blank()))
(p_nit  <- plot_diel_with_sig("NOx",  expression(paste("NO"[x], " (", mu, "mol L"^-1,")")), free_y = TRUE) + theme(strip.text.x = element_blank(), strip.background.x = element_blank()))
(p_nh4 <- plot_diel_with_sig("NH4",    expression(paste("NH"[4]," (µmol L"^-1,")" ))) + theme(strip.text = element_blank(), strip.background = element_blank()))
(p_po4 <- plot_diel_with_sig("PO4",  expression(paste("Phosphate (µmol L"^-1,")"))) + theme(strip.text.x = element_blank(), strip.background.x = element_blank()))
(p_mc <- plot_diel_with_sig("M_to_C", "M:C")
  + theme(strip.text = element_blank(), strip.background = element_blank()))
(p_bix  <- plot_diel_with_sig("BIX",  "Biological Index (BIX)")
  + theme(strip.text.x = element_blank(), strip.background.x = element_blank()))
(p_hix  <- plot_diel_with_sig("HIX",  "Humification Index (HIX)")
  + theme(strip.text = element_blank(), strip.background = element_blank()))


p_ph + p_TA + p_nit + p_nh4+ p_bix + p_hix + 
  plot_layout(axes = "collect_x", ncol = 2)

#ggsave(here("Output","DielPlots.png"), width = 14, height = 14) 
```

## Effect size plots
```{r effect-size-plot, fig.height = 10}

build_diel_effect_df_one_scaled <- function(var_name, data = data_clean) {
  stopifnot(var_name %in% names(data))
  
  # scale the variable (global z-score)
  data_z <- data %>%
    mutate(var_z = as.numeric(scale(.data[[var_name]])))
  
  fml <- as.formula("var_z ~ Day_Night * Season * Benthos + (1|Quad_ID)")
  fit <- lmer(fml, data = data_z)
  
  # emmeans for Day_Night within each Season × Benthos
  emm <- emmeans(fit, ~ Day_Night | Season * Benthos)
  
  # Dusk - Dawn contrast in *scaled units* (≈ SD units)
  diel <- contrast(emm, method = "revpairwise")
  
  # summary with p-values + 95% CI
  df <- summary(diel, infer = c(TRUE, TRUE), level = 0.95) %>%
    as.data.frame()
  
  # standardize CI names
  if (!"lower.CL" %in% names(df)) {
    if ("asymp.LCL" %in% names(df)) {
      df <- df %>%
        rename(
          lower.CL = asymp.LCL,
          upper.CL = asymp.UCL
        )
    } else if ("LCL" %in% names(df)) {
      df <- df %>%
        rename(
          lower.CL = LCL,
          upper.CL = UCL
        )
    }
  }
  
  df %>%
    mutate(
      est_std   = estimate,     
      lower_std = lower.CL,
      upper_std = upper.CL,
      variable  = var_name
    )
}

effects_all <- map_dfr(
  less_vars,
  build_diel_effect_df_one_scaled,
  data = data_clean
)

effects_all <- effects_all %>%
  mutate(
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ "ns"
    ),
    variable = factor(variable, levels = rev(less_vars))
  )

effectsizeplot <-
ggplot(effects_all,
       aes(x = est_std, y = variable,
           color = Season,  
           alpha = sig != "ns")) +   
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey70") +
  geom_errorbarh(aes(xmin = lower_std, xmax = upper_std),
                 height = 0.15) +
  geom_point(size = 2) +
  facet_wrap(~ Benthos, ncol = 5) +
  scale_y_discrete(labels = c("M_to_C" = "M:C"))+ # adjust M:C format
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.3), # make sig darker
                     guide = "none") +
  scale_color_manual(values = c(
    "Fall"   = "#D95F02",
    "Spring" = "#1B9E77"
  )) +
  labs(
    x = "Standardized effect size (Dusk – Dawn, SD units)",
    y = NULL,
    color = "Season"
  ) +
  
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    axis.text.y = element_text(size = 9)
  )
```

### Patchworked plots
```{r effect-box, fig.height = 8}
(p_ph   <- plot_diel_with_sig("pH",   "pH (Total Scale)"))
(p_nit  <- plot_diel_with_sig("NOx",  expression(paste("NO"[x], " (", mu, "mol L"^-1,")")), free_y = TRUE) + theme(strip.text.x = element_blank(), strip.background.x = element_blank()))
(p_hix  <- plot_diel_with_sig("HIX",  "Humification Index (HIX)")+
  theme(strip.text.x = element_blank(), strip.background.x = element_blank()))

effectsizeplot | free(p_ph / p_nit / p_hix + plot_layout(axes = "collect"))

design <- "###BBB
           AAABBB
           AAABBB
           AAABBB
           AAABBB
           ###BBB"

effectsizeplot + (p_ph / p_nit / p_hix + plot_layout(axes = "collect")) + plot_layout(design = design)

(p_ph   <- plot_diel_with_sig("pH",   "pH (Total Scale)"))
(p_nit  <- plot_diel_with_sig("NOx",  expression(paste("NO"[x], " (", mu, "mol L"^-1,")")), free_y = TRUE))
(p_hix  <- plot_diel_with_sig("HIX",  "Humification Index (HIX)"))

effectsizeplot + free(p_ph) + p_nit + p_hix + plot_layout(axes = "collect")
```

# Heatmap colored in comparison to OO
Calculates means per benthos x season x day/night and compares against the Surface Water mean
```{r heatmap-plot-code}
# take means of all variables
heatmap_data <- data_clean %>%
  group_by(Benthos, Day_Night, Season) %>%
  summarise(across(all_of(less_vars), ~ mean(.x, na.rm = TRUE)),
            .groups = "drop") %>%
  pivot_longer(cols = all_of(less_vars),
               names_to = "variable",
               values_to = "Mean_Value")

# get Surface Water means for comparison
open_ocean_ref <- heatmap_data %>%
  filter(Benthos == "Surface Water") %>%
  select(-Benthos) %>%
  rename(OpenOcean_Value = Mean_Value)

# join both to calculate ΔOO
fdom_diff <- heatmap_data %>%
  filter(Benthos != "Surface Water") %>% 
  left_join(open_ocean_ref,
            by = c("Day_Night","Season","variable")) %>%
  mutate(Diff_from_OO = Mean_Value - OpenOcean_Value)

# scale the differences so we can plot it
fdom_diff <- fdom_diff %>%
  group_by(variable) %>%
  mutate(Diff_Scaled = as.numeric(scale(Diff_from_OO)), # scale function scales AND centers, leading to inconsistencies
         Diff_Scaled_Nocenter = as.numeric(scale(Diff_from_OO, center = FALSE))) %>%  # scales without centering
  ungroup() %>%
  mutate(Benthos = factor(Benthos, 
    levels = c("Mussel", "Barnacle", "Surfgrass", "Rockweed"))) # order benthos by assigning levels

# assign var labels
var_levels <- c("pH", "NOx", "NH4", "HIX")

plot_df <- fdom_diff %>%
  mutate(
    variable   = factor(variable, levels = var_levels),
    Day_Night  = factor(Day_Night, levels = c("Dawn", "Dusk")),
    x_id       = as.numeric(variable) * 2 + as.numeric(Day_Night) - 2
    # give an x value to each variable in order
  )

# positions for bottom labels (every column)
bottom_breaks  <- sort(unique(plot_df$x_id))

# positions for top labels (center of each pair)
top_breaks     <- seq(1.5, by = 2, length.out = length(var_levels))

# vertical separators between variables
sep_positions  <- seq(2.5, by = 2, length.out = length(var_levels) - 1)

# plot
ggplot(plot_df, aes(x = x_id, y = Benthos, fill = Diff_Scaled_Nocenter)) +
  geom_tile(color = "white") +
  geom_vline(xintercept = sep_positions, color = "white", linewidth = 1.5) +
  scale_fill_gradient2(
    low = "#075AFF", mid = "white", high = "#FF4500",
    midpoint = 0, name = "Scaled difference\nfrom surface water"
  ) +
  scale_x_continuous(
    breaks = bottom_breaks,
    labels = rep(c("Dawn", "Dusk"), times = length(var_levels)),
    sec.axis = sec_axis(
      ~ .,
      breaks = top_breaks,
      labels = var_levels,
      name = NULL
    )
  ) +
  labs(
    title = "Deviation from Surface Water fDOM + Nutrients",
    x = NULL,
    y = NULL
  ) +
  facet_grid(Season ~ .) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 0),
    axis.text.x.top = element_text(size = 10),
    axis.ticks.x.top = element_blank(),
    strip.text = element_text(size = 10)
  )
```

```{r heatmap-ANOVA, eval=FALSE}
# filter out NAs
testing_df_norm <- testing_df_norm %>% 
  filter(!is.na(Value))%>%
  mutate(Season = as.factor(Season))

compute_sig_to_ctrl <- function(testing_df_norm, vars = numeric_vars_normal) {
  
  options(contrasts = c("contr.sum", "contr.poly")) # set contrasts for type III ANOVA
  
  res_tbl <- map(vars, function(v) {
    df <- testing_df_norm %>%
      filter(variable == v)
    
    fit <- lm(Value ~ Benthos * Day_Night * Season, data = df) # create lm
    model <- Anova(fit, type = "III")
    emm <- emmeans(fit, ~ Benthos | Day_Night * Season) # emmeans for tukey's hsd
    vs_ctrl <- contrast(emm, method = "trt.vs.ctrl", ref = control_label, adjust = "mvt")
    td  <- as.data.frame(vs_ctrl)
    
    comp_tbl <- td %>%
      separate(contrast, c("left","right"), sep = " - ", remove = FALSE) %>%
      mutate(
        Benthos  = if_else(right == control_label, left, right),
        variable = v,
        sig = case_when(
          is.na(p.value)  ~ NA_character_,
          p.value < 0.001 ~ "***",
          p.value < 0.01  ~ "**",
          p.value < 0.05  ~ "*",
          TRUE            ~ ""
        )
      ) 
    ctrl_tbl <- crossing(
      variable = v,
      Season   = unique(td$Season),
      Day_Night= unique(td$Day_Night),
      Benthos  = control_label
    ) %>%
      mutate(
        estimate = NA_real_, SE = NA_real_, df = NA_real_, t.ratio = NA_real_,
        p.value  = NA_real_, sig = ""
      )

    bind_rows(comp_tbl, ctrl_tbl) %>%
      distinct(variable, Season, Day_Night, Benthos, .keep_all = TRUE)
  }) %>% compact() %>% list_rbind()

  res_tbl
}

sig_to_ctrl <- compute_sig_to_ctrl(testing_df_norm) %>%
  mutate(variable = sub("_transformed$", "", variable)) # convert name back for join


# get n so we can add to the plot
n_table <- testing_df_norm %>%
  mutate(variable = sub("_transformed$", "", variable)) %>%
  group_by(Season, Benthos, variable, Day_Night) %>%
  summarise(n = sum(!is.na(Value)), .groups = "drop")

# join all data for plotting
plot_df_anova <- fdom_diff %>%  
  left_join(sig_to_ctrl, by = c("Season","Day_Night","Benthos","variable")) %>%
  left_join(n_table,   by = c("Season","Day_Night","Benthos","variable")) %>%
  mutate(variable   = factor(variable, levels = var_levels),
         Day_Night  = factor(Day_Night, levels = c("Dawn", "Dusk")),
         x_id       = as.numeric(variable) * 2 + as.numeric(Day_Night) - 2)
  
# updated plot
ggplot(plot_df_anova, aes(x = x_id, y = Benthos, fill = Diff_Scaled_Nocenter)) +
  geom_tile(color = "white") +
  geom_vline(xintercept = sep_positions, color = "white", linewidth = 1.5) +
  scale_fill_gradient2(
    low = "#075AFF", mid = "white", high = "#FF4500",
    midpoint = 0, name = "Scaled difference\nfrom surface water"
  ) +
  scale_x_continuous(
    breaks = bottom_breaks,
    labels = rep(c("Dawn", "Dusk"), times = length(var_levels)),
    sec.axis = sec_axis(
      ~ .,
      breaks = top_breaks,
      labels = var_levels,
      name = NULL
    )
  ) +
  labs(
    title = "Dusk vs Dawn fDOM + Nutrients (orig ANOVA)",
    x = NULL,
    y = NULL
  ) +
  facet_grid(Season ~ .) +
  geom_text(aes(label = ifelse(is.na(sig),"", paste0(sig))), color = "black", size = 3) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 0),
    axis.text.x.top = element_text(size = 10),
    axis.ticks.x.top = element_blank(),
    strip.text = element_text(size = 10)
  )
```

## One-sample t-test
```{r heatmap-ttest, fig.width = 8}
compute_sig_to_ctrl_ttest <- function(testing_df_norm, 
                                      vars = numeric_vars_normal,
                                      control_label) {
  
  res_tbl <- map(vars, function(v) { # loop for each variable
    df_var <- testing_df_norm %>%
      filter(variable == v)
    
    res_by_sd <- df_var %>%
      group_nest(Season, Day_Night) %>% # group by season and dawn/dusk
      mutate(
        res = map(data, function(df_sd) {
          ctrl_vals <- df_sd %>%
            filter(Benthos == control_label) %>%
            pull(Value)
          ctrl_mean <- mean(ctrl_vals, na.rm = TRUE) # get OO mean

          df_sd %>%
            group_by(Benthos) %>%
            summarise(
              n       = sum(!is.na(Value)),
              mean    = mean(Value, na.rm = TRUE),
              .groups = "drop"
            ) %>%
            mutate(
              # one-sample test vs control mean for each benthos
              p.value = map_dbl(Benthos, function(b) {
                x <- df_sd %>% filter(Benthos == b) %>% pull(Value)
                t.test(x, mu = ctrl_mean)$p.value
              }),
              estimate = mean - ctrl_mean,
              variable = v
            )
        })
      ) %>%
      select(-data) %>%
      unnest(res)
    
    # add sig stars
    res_by_sd %>%
      mutate(
        sig = case_when(
          is.na(p.value)  ~ NA_character_,
          p.value < 0.001 ~ "***",
          p.value < 0.01  ~ "**",
          p.value < 0.05  ~ "*",
          TRUE            ~ ""
        )
      )
  }) %>% 
    compact() %>% 
    list_rbind()
  
  res_tbl
}

sig_to_ctrl_ttest <- compute_sig_to_ctrl_ttest(testing_df_norm,
                                               vars = numeric_vars_normal,
                                               control_label = "Surface Water")%>%
  mutate(variable = sub("_transformed$", "", variable))

plot_df_ttest <- fdom_diff %>%
  left_join(sig_to_ctrl_ttest, 
            by = c("Season","Day_Night","Benthos","variable")) %>%
  mutate(variable   = factor(variable, levels = var_levels),
         Day_Night  = factor(Day_Night, levels = c("Dawn", "Dusk")),
         x_id       = as.numeric(variable) * 2 + as.numeric(Day_Night) - 2,
         Benthos    = factor(Benthos, levels = c("Mussel","Barnacle", "Surfgrass", "Rockweed")))

# updated plot
ggplot(plot_df_ttest, aes(x = x_id, y = Benthos, fill = Diff_Scaled_Nocenter)) +
  geom_tile(color = "white") +
  geom_vline(xintercept = sep_positions, color = "white", linewidth = 1.5) +
  scale_fill_gradient2(
    low = "#075AFF", mid = "white", high = "#FF4500",
    midpoint = 0, name = "Scaled difference\nfrom surface water"
  ) +
  scale_x_continuous(
    breaks = bottom_breaks,
    labels = rep(c("Dawn", "Dusk"), times = length(var_levels)),
    sec.axis = sec_axis(
      ~ .,
      breaks = top_breaks,
      labels = var_levels,
      name = NULL
    )
  ) +
  labs(
    x = NULL,
    y = NULL
  ) +
  facet_grid(Season ~ .) +
  geom_text(aes(label = ifelse(is.na(sig),"", paste0(sig))), color = "black", size = 3) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 0),
    axis.text.x.top = element_text(size = 10),
    axis.ticks.x.top = element_blank(),
    strip.text = element_text(size = 10)
  )

```

```{r heatmap-dunn, eval=FALSE}
compute_sig_to_ctrl_dunn <- function(testing_df_norm, 
                                     vars = numeric_vars,
                                     control_label) {
  
  res_tbl <- map(vars, function(v) { # loop for each variable
    df_var <- testing_df_norm %>%
      filter(variable == v)
    
    df_var %>%
      group_by(Season, Day_Night) %>% # group by season and dawn/dusk
      group_modify(~ {
        df_sd <- .x
  
        dunn_res <- df_sd %>%
          rstatix::dunn_test(Value ~ Benthos, p.adjust.method = "none")
        
        # keep only comparisons involving OO
        dunn_ctrl <- dunn_res %>%
          filter(group1 == control_label | group2 == control_label) %>%
          mutate(
            Benthos  = if_else(group1 == control_label, group2, group1),
            variable = v,
            p.value  = p.adj
          )
        
        dunn_ctrl
      }) %>%
      ungroup()
  }) %>%
    compact() %>%
    list_rbind() %>%
    mutate(
      sig = case_when(
        is.na(p.value)  ~ NA_character_,
        p.value < 0.001 ~ "***",
        p.value < 0.01  ~ "**",
        p.value < 0.05  ~ "*",
        TRUE            ~ ""
      )
    )
  
  res_tbl
}

sig_to_ctrl_dunn <- compute_sig_to_ctrl_dunn(testing_df_norm,
                                             vars = numeric_vars,
                                             control_label = control_label) %>%
  mutate(variable = sub("_transformed$", "", variable))

plot_df_dunn <- fdom_diff %>%
  left_join(sig_to_ctrl_dunn, 
            by = c("Season","Day_Night","Benthos","variable")) %>%
  left_join(n_table, 
            by = c("Season","Day_Night","Benthos","variable")) %>%
  mutate(variable   = factor(variable, levels = var_levels),
         Day_Night  = factor(Day_Night, levels = c("Dawn", "Dusk")),
         x_id       = as.numeric(variable) * 2 + as.numeric(Day_Night) - 2)

# create plot
ggplot(plot_df_dunn, aes(x = x_id, y = Benthos, fill = Diff_Scaled_Nocenter)) +
  geom_tile(color = "white") +
  geom_vline(xintercept = sep_positions, color = "white", linewidth = 1.5) +
  scale_fill_gradient2(
    low = "#075AFF", mid = "white", high = "#FF4500",
    midpoint = 0, name = "Scaled difference\nfrom surface water"
  ) +
  scale_x_continuous(
    breaks = bottom_breaks,
    labels = rep(c("Dawn", "Dusk"), times = length(var_levels)),
    sec.axis = sec_axis(
      ~ .,
      breaks = top_breaks,
      labels = var_levels,
      name = NULL
    )
  ) +
  labs(
    title = "Dusk vs Dawn fDOM + Nutrients (Dunn's test)",
    x = NULL,
    y = NULL
  ) +
  facet_grid(Season ~ .) +
  geom_text(aes(label = ifelse(is.na(sig),"", paste0(sig))), color = "black", size = 3) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 0),
    axis.text.x.top = element_text(size = 10),
    axis.ticks.x.top = element_blank(),
    strip.text = element_text(size = 10)
  )

```

# Tidal exchange
Samples were all collected at "high tide," but does it matter if water is coming in (flood) or leaving (ebb)? 
```{r ebb-flood}
# what is our spread of tidal exchange over dusk/dawn?
ggplot(testing_df_norm %>%
         mutate(Tide = factor(Tide, levels = c("ebb", "flood"),
                              labels = c("Ebb", "Flood"))), 
       aes(Tide, fill = Day_Night)) +
  facet_wrap(~Season)+
  geom_bar(position = "fill")+
  scale_fill_manual(values = c("Dawn" = "#FDB813", "Dusk" = "#003865")) +
  labs(
    x = "Tidal exchange",
    y = "Proportion of samples",
    fill = "Sampling Time"
  )

ggplot(testing_df_norm %>%
         mutate(Tide = factor(Tide, levels = c("ebb", "flood"),
                              labels = c("Ebb", "Flood"))), 
         aes(Day_Night, fill = Tide)) +
  facet_wrap(~Season)+
  geom_bar(position = "fill")+
  scale_fill_manual(values = c("Flood" = "#C28FFF", "Ebb" = "#A97C49")) +
  labs(
    x = "Sampling time",
    y = "Proportion of samples",
    fill = "Tidal Exchange"
  )

# NOx plot
ggplot(testing_df %>% filter(Season == "Spring", variable == "NOx"), 
       aes(x = Day_Night, y = Value, fill = Day_Night))+
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
  scale_fill_manual(values = c("Dawn" = "#FDB813", "Dusk" = "#003865")) +
  facet_grid(rows = vars(Tide), cols = vars(Benthos))+
  labs(x = "Sampling Time",
       y = "NOx")

# no dawn ebb samples during fall, and few dusk ebb samples during spring

# okay, well does there seem to be an effect regardless of season?
anovas_full_ebb <- map_dfr(numeric_vars_normal, function(v){
  df_sub <- testing_df_norm %>%
    filter(variable == v) # filter by variable
    options(contrasts = c("contr.sum", "contr.poly"))
    fit <- lm(Value ~ Day_Night * Benthos + Tide, data = df_sub)
    Anova(fit, type = "III") %>%
    as.data.frame() %>%
    rownames_to_column("term") %>%
    filter(term != "(Intercept)") %>%
    filter(`Pr(>F)` < 0.05) %>% # see what's significant
    mutate(variable = v)
})
anovas_full_ebb %>%
  kable()
# potentially for pH and HIX


# lets visualize ebb vs flood in fall at dusk & in spring at dawn and see
ggplot(testing_df_norm %>%
         filter(Season == "Fall",
                Day_Night == "Dusk",
                variable == "pH"), 
       aes(x = Tide, y = Value)) +
  geom_boxplot()

ggplot(testing_df_norm %>%
         filter(Season == "Spring",
                Day_Night == "Dawn",
                variable == "pH"), 
       aes(x = Tide, y = Value)) +
  geom_boxplot()

# overall makes sense for water coming in (flood) to be higher pH, maybe see a lower effect of this in Spring because of upwelling

# lets look at all numeric vars just in case

# fall dusk = pH and humics
anovas_full_tidal_fall <- map_dfr(numeric_vars_normal, function(v){
  df_sub <- testing_df_norm %>%
    filter(variable == v,
           Season == "Fall",
          Day_Night == "Dusk")
    options(contrasts = c("contr.sum", "contr.poly"))
    fit <- lm(Value ~ Tide, data = df_sub)
    Anova(fit, type = "III") %>%
    as.data.frame() %>%
    rownames_to_column("term") %>%
    filter(term != "(Intercept)") %>%
    filter(`Pr(>F)` < 0.05) %>% # see what's significant
    mutate(variable = v)
})

# spring nothing
anovas_full_tidal_spring <- map_dfr(numeric_vars_normal, function(v){
  df_sub <- testing_df_norm %>%
    filter(variable == v,
           Season == "Spring",
          Day_Night == "Dawn")
    options(contrasts = c("contr.sum", "contr.poly"))
    fit <- lm(Value ~ Tide, data = df_sub)
    Anova(fit, type = "III") %>%
    as.data.frame() %>%
    rownames_to_column("term") %>%
    filter(term != "(Intercept)") %>%
    filter(`Pr(>F)` < 0.05) %>% # see what's significant
    mutate(variable = v)
})

```


# Community Composition

## Sessile
```{r sessile-barplot}
## sum across the canopy and understory ###

sessile_total <-sessile %>%
  group_by(plot, season)%>%
  summarise_at(.vars = vars(Silvetia_compressa:Chondracanthus_canaliculatus), .funs = sum)%>% 
  mutate(Total_hairs = rowSums(across(Silvetia_compressa:Chondracanthus_canaliculatus))) %>% # count total crosshairs
  ungroup()%>% # calculate percent cover
  mutate(across(c(Silvetia_compressa:Chondracanthus_canaliculatus),  ~ . /Total_hairs*100))%>%
  rename(BenthicID = plot, Season = season)%>% # keep consistent with other data
  mutate(Season = factor(Season, levels = c("summer2023","fall2023","winter2024", "spring2024")))%>%
  mutate(Benthos = case_when( # add benthos name
    grepl("B", BenthicID) ~ "Barnacles",
    grepl("M", BenthicID) ~"Mussels",
    grepl("P", BenthicID) ~"Surfgrass",
    grepl("R", BenthicID) ~"Rockweed"), 
    .after = BenthicID)


## Make a stacked barplot of all the species across seasons
sessile_total %>%
  group_by(Benthos, Season)%>% # calculate the average for each season
  summarise(across(c(Silvetia_compressa:Chondracanthus_canaliculatus),  ~mean(.,na.rm = TRUE))) %>%
  pivot_longer(cols = c(Silvetia_compressa:Chondracanthus_canaliculatus), names_to = "Species", values_to = "Percent_Cover")%>%
  left_join(sessile_function)%>%
  filter(Season %in% c('fall2023','spring2024')) %>%
  mutate(Season = factor(Season, 
    levels =  c("fall2023", "spring2024"), 
    labels = c("Fall", "Spring"))) %>%
  ggplot(aes(x = Season,y = Percent_Cover, fill = FunctionalGroup ))+
    geom_bar(position="stack", stat="identity")+
  scale_fill_paletteer_d("MetBrewer::Signac")+
  labs(x = "",
       y = "% Cover",
       color = "",
       fill = "Functional Group")+
  theme_bw()+
  facet_wrap(~Benthos)

#ggsave(here("Output","SessileComp.png"), width = 10, height = 6)
```

## Mobiles
```{r mobile-barplot}

## calculate denisity per m2 (used 0.25m2 quad)
mobile_density <-
  mobile %>%
  rename(BenthicID = plot, Season = season)%>% # keep consistent with other data
  mutate(Season = factor(Season, levels = c("summer2023","fall2023","winter2024","spring2024")))%>%
  mutate(Benthos = case_when( # add benthos name
    grepl("B", BenthicID) ~ "Barnacles",
    grepl("M", BenthicID) ~"Mussels",
    grepl("P", BenthicID) ~"Surfgrass",
    grepl("R", BenthicID) ~"Rockweed"), 
    .after = BenthicID)%>%
  select(-c(Littorina_sp_raw, Lottia_sp_raw))%>%
  mutate(across(c(Lottia_sp_scaled:Strongylocentrotus_purpuratus),  ~./0.25))
  

mobile_density %>%
  group_by(Benthos, Season)%>% # calculate the average for each season
  summarise(across(c(Lottia_sp_scaled:Strongylocentrotus_purpuratus),  ~mean(.,na.rm = TRUE))) %>%
  pivot_longer(cols = c(Lottia_sp_scaled:Strongylocentrotus_purpuratus), names_to = "Species", values_to = "Density_m2")%>%
  left_join(mobile_function)%>% #join with functional data
  filter(Season %in% c('fall2023','spring2024')) %>%
  mutate(Season = factor(Season, 
    levels =  c("fall2023", "spring2024"), 
    labels = c("Fall", "Spring"))) %>%
  ggplot(aes(x = Season,y = Density_m2, fill = FunctionalGroup ))+
  geom_bar(position="stack", stat="identity")+
  scale_fill_paletteer_d("MetBrewer::Signac")+
  labs(x = "",
       y = expression(paste("Mean Density (m"^2,")")),
       color = "",
       fill = "Functional Group")+
  theme_bw()+
  facet_wrap(~Benthos, scale = "free")

#ggsave(here("Output","MobileComp.png"), width = 10, height = 6)

```