---
title: "intertidal_analysis_short"
author: "tena"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
  github_document:
    html_preview: false
params:
  keep_times: ["Dawn", "Dusk"]
  control_label: "Open Ocean"
---

# Overview

This document:

- Loads and merges pH/TA/nutrients and fDOM datasets
- Cleans data and applies outlier handling (IQR rule)
- Summarizes by habitat and diel period
- Creates PCA for seasonal drivers
- Visualizes diel and habitat patterns (boxplots, TA–pH, fDOM heatmap)
- Computes correlation matrices and significance masks
- Runs ANOVAs
- Visualizes ANOVA results and reassesses hypotheses

> **Reproducibility & Modularity**  
> Kept times (which removes summer), and the control label (open ocean) are parameterized in the YAML `params:` so I can change inputs without editing code chunks.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(patchwork)
library(rstatix)
library(ggcorrplot)
library(ggrepel)
library(here)
library(lubridate)
library(kableExtra)
library(car)
library(emmeans)
library(agricolae)
library(lme4)
library(paletteer)

# Global plot theme
theme_set(theme_bw(base_size = 12))
```

## Paths & Parameters

```{r params}
keep_times <- params$keep_times
control_label <- params$control_label
```

# Data Loading & Merging

This section loads both fDOM and pH/nutrients CSVs, parses dates, and joins on shared identifiers. We keep all samples, regarless of missing data (`full_join`).

```{r data_load}
load_ok <- TRUE
tryCatch({
  data_ph <- read_csv(here::here("Data","Biogeochemistry","pHProbe_Data_calculated.csv"), na = c("NA", ""), show_col_types = FALSE) 
    #mutate(Sampling_Date = lubridate::mdy(Sampling_Date))
  data_fdom <- read_csv(here::here("Data","Biogeochemistry","fDOM_clean.csv"), na = c("NA", ""), show_col_types = FALSE) %>%
    mutate(Sampling_Date = lubridate::mdy(Sampling_Date))

  data_merged <- full_join(
    data_ph,
    data_fdom,
    by = c("UniqueID", "Benthos", "Quad_ID", "Day_Night", "Sampling_Date", "Sampling_Time","Season")
  ) %>%
    mutate(Humic = UVHumic+VisibleHumic+MarineHumic,# make a column for humics
           Prot = Tryptophan+Tyrosine, # make a column for proteinaceaus fDOM
           Nitrogen = NO3+NO2 # make a column of total N
           ) 

  cat("Data loaded and merged successfully.\n")
  cat("Total merged samples:", nrow(data_merged), "\n")
}, error = function(e){
  load_ok <<- FALSE
  message("Error loading or merging files. Please check file paths and names.")
  message(e)
})

sessile <- read_csv(here("Data","Community_Composition","sessile_data.csv"))
sessile_function <-read_csv(here("Data","Community_Composition","SessileFunctional.csv"))

mobile <- read_csv(here("Data","Community_Composition","mobile_data.csv"))
mobile_function <-read_csv(here("Data","Community_Composition","MobileFunctional.csv"))

stopifnot(load_ok)
```

# Cleaning & Pre-processing

We declare variable sets for identifiers, water chemistry, nutrients, and fDOM components. We retain only **Dawn** and **Dusk** samples, coerce numeric columns, set factor orders, drop NA grouping levels, and apply the **1.5×IQR rule** as a conservative outlier filter (replacing outliers with `NA`).

```{r clean-variables}

# Identifier variables
id_vars <- c("UniqueID", "Season", "Benthos", "Quad_ID", "Day_Night", "Sampling_Date")

# Water chemistry variables (from pH probe file)
water_chem_vars <- c("Salinity", "pH", "TempInSitu", "TA")

# Nutrient variables (from pH probe file)
nutrient_vars <- c("NO3", "PO4", "SIL", "NO2", "NH4", "Nitrogen")

# fDOM variables (from fDOM file)
#fdom_vars <- c("UVHumic", "Tyrosine", "VisibleHumic", "MarineHumic",
#               "Tryptophan", "FulvicAcid", "BIX", "HIX", "FI", "M_to_C")

# reduced variables
fdom_vars <- c("Humic","Prot", "BIX", "HIX", "M_to_C")

all_numeric_vars <- c(water_chem_vars, nutrient_vars, fdom_vars)

# only vars of interest
numeric_vars <- c("pH", "TA", "Nitrogen", "Humic", "Prot", "BIX", "HIX", "M_to_C")

# Helper to IQR-filter a numeric vector (returning NA for outliers)
iqr_filter <- function(x){
  if(!is.numeric(x)) return(x)
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lo <- q1 - 1.5 * iqr
  hi <- q3 + 1.5 * iqr
  ifelse(x < lo | x > hi, NA_real_, x)
}

# Cleaned dataset
data_clean <- data_merged %>%
  filter(Day_Night %in% keep_times) %>%
  select(-M_to_C) %>%
  mutate(M_to_C = MarineHumic / VisibleHumic) %>%
  select(all_of(id_vars), any_of(all_numeric_vars)) %>%
  mutate(across(all_of(all_numeric_vars), as.numeric)) %>%
  mutate(
    Benthos   = factor(Benthos, levels = c(control_label, "Rockweed", "Surfgrass", "Barnacle", "Mussel")),
    Day_Night = factor(Day_Night, levels = c("Dawn", "Dusk"))
  ) %>%
  drop_na(Benthos, Day_Night) %>%
  
  mutate(across(all_of(all_numeric_vars), iqr_filter))
```

# Summary Statistics

Per habitat × diel period (and Season), we compute mean, SD, and sample size for all numeric variables.

```{r summary-stats}

data_summary <- data_clean %>%
  group_by(Benthos, Day_Night, Season) %>%
  summarise(
    across(all_of(all_numeric_vars), list(mean = ~ mean(., na.rm = TRUE),
                                          sd   = ~ sd(., na.rm = TRUE),
                                          n    = ~ sum(!is.na(.))
    )),
    .groups = "drop"
  )

data_summary %>%
  kable()
```

# Background plots

## Correlation Matrix

We compute a Pearson correlation matrix on complete cases and visualize it with significance masking.
```{r corr-matrix}
cor_data <- data_clean %>% select(all_of(numeric_vars)) %>% drop_na()

if(nrow(cor_data) > 0){
  cor_mat <- rstatix::cor_mat(cor_data)
  cor_p   <- rstatix::cor_pmat(cor_data)

  ggcorrplot(
    cor_mat,
    hc.order = TRUE,
    type = "lower",
    p.mat = cor_p,
    sig.level = 0.05,
    insig = "blank",
    lab = TRUE,
    lab_size = 2.5,
    colors = c("#6D9EC1", "white", "#E46726"),
    title = "Correlation Matrix of Biogeochemical Variables\n(white not significant)"
  )
} else {
  cat("Not enough complete data (after drop_na) to compute correlation matrix or PCA.\n")
}
```

### PCA plots
```{r PCA-func, echo=FALSE}
pca_biplot <- function(df,
                       vars,
                       title = "PCA biplot",
                       color_var = "Season",
                       shape_var = "Day_Night",
                       facet_var = "Benthos") {
  
  # Keep only needed columns
  df_sub <- df %>%
    dplyr::select(all_of(c(vars, color_var, shape_var, facet_var))) %>%
    tidyr::drop_na(all_of(vars))
  
  # Numeric matrix for PCA
  num_mat <- df_sub %>%
    dplyr::select(all_of(vars)) %>%
    as.matrix()
  
  # Run PCA (scaled)
  pca_res <- prcomp(num_mat, scale. = TRUE)
  
  # Scores (samples)
  scores <- as.data.frame(pca_res$x[, 1:2])
  colnames(scores) <- c("PC1", "PC2")
  scores <- cbind(scores,
                  df_sub %>% dplyr::select(all_of(c(color_var, shape_var, facet_var))))
  
  # Loadings (variable arrows)
  loadings <- as.data.frame(pca_res$rotation[, 1:2])
  colnames(loadings) <- c("PC1", "PC2")
  loadings$var <- rownames(loadings)
  
  # Scale arrows to fit in score space
  mult <- min(
    (max(scores$PC1) - min(scores$PC1)) / (max(loadings$PC1) - min(loadings$PC1)),
    (max(scores$PC2) - min(scores$PC2)) / (max(loadings$PC2) - min(loadings$PC2))
  ) * 0.7
  
  loadings_scaled <- loadings
  loadings_scaled$PC1 <- loadings_scaled$PC1 * mult
  loadings_scaled$PC2 <- loadings_scaled$PC2 * mult
  
  # % variance explained
  var_expl <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100
  
  ggplot(scores, aes(x = PC1, y = PC2)) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    geom_point(aes_string(color = color_var, shape = shape_var),
               size = 3, alpha = 0.8) +
    geom_segment(data = loadings_scaled,
                 aes(x = 0, y = 0, xend = PC1, yend = PC2),
                 arrow = arrow(length = unit(0.2, "cm"))) + #arrows
    geom_text_repel(
    data = loadings_scaled,
    aes(x = PC1, y = PC2, label = var),
    size = 4,
    segment.color = "grey50",
    max.overlaps = Inf
)+
    facet_wrap(as.formula(paste("~", facet_var)))+
    scale_color_paletteer_d("nationalparkcolors::Arches")+
    #scale_color_manual(values = c("Fall" = "#bf2624", "Spring" = "#6dd8bf"))+
    stat_ellipse(aes_string(color=color_var), level = 0.95)+ #circles
    labs(
      title = title,
      x = paste0("PC1 (", round(var_expl[1], 1), "%)"),
      y = paste0("PC2 (", round(var_expl[2], 1), "%)")
    ) +
    theme_bw()
}

```


## Seasonal differences - PCA

Do different seasons have distinct chemical/fDOM signatures? Which variables drive those differences?

```{r PCA-plots, fig.height=8}
# original one without benthos (unfaceted)
pca_biplot(
  data_clean,
  vars      = numeric_vars,
  title     = "PCA of Nutrients and fDOM components driving seasonal differences",
  color_var = "Season",
  shape_var = "Day_Night",
  facet_var = "Season"
)

p_pca_season <- pca_biplot(
  data_clean,
  vars      = numeric_vars,
  title     = "PCA of Nutrients and fDOM components driving seasonal differences",
  color_var = "Season",
  shape_var = "Day_Night",
  facet_var = "Benthos"
)

p_pca_season

p_pca_benthos <- pca_biplot(
  data_clean,
  vars = numeric_vars,
  title = "PCA of Nutrients and fDOM components across benthos by season",
  color_var = "Benthos",
  shape_var = "Day_Night",
  facet_var = "Season"
)

p_pca_benthos

p_pca_diel <- pca_biplot(
  data_clean,
  vars = numeric_vars,
  title = "PCA of nutrients and fDOM components across benthos by dawn/dusk",
  color_var = "Benthos",
  shape_var = "Season",
  facet_var = "Day_Night"
)

p_pca_diel
```

Spring shows strong nutrient and DOM enrichment, can also see diel variability in nutrient accumulation.

# Dawn vs Dusk analyses across Benthos and Seasons

This could be due to lost samples using diel variability, let's look at Day_Night results

```{r ANOVA, fig.keep = "none"}
# create long format df
testing_df <- data_clean %>%
  pivot_longer(cols = all_of(all_numeric_vars),
               names_to = "variable", values_to = "Value")

nutrient_vars <- c("NO3", "PO4", "SIL", "NO2", "NH4", "Nitrogen")
all_numeric_vars_normal <- c(water_chem_vars, nutrient_vars, fdom_vars)

# running lm and creating diagnostic plots
map_dfr(all_numeric_vars, function(v){
  df_sub <- testing_df %>%
    filter(variable == v) # filter by variable
    options(contrasts = c("contr.sum", "contr.poly"))
    fit <- lm(Value ~ Day_Night * Benthos * Season, data = df_sub)
  par(mfrow = c(2,2)) # create 2x2 for plots
  plot(fit, main = paste("Diagnostics for", v)) # check assumptions using qq and residuals plots
})

# okay NO3, PO4, Humic, Prot, HIX looks like they need to be transformed

transform_var <- c("NO3", "PO4", "Humic", "Prot", "HIX")
transformed_var <- map_dfr(transform_var, function(v){
  testing_df %>%
    filter(variable == v) %>% # filter by variable
    mutate(Value = sqrt(Value),
           variable = paste0(v,"_transformed"))
})

```

just the anova results

```{r ANOVA-table}
testing_df_norm <- bind_rows(testing_df, transformed_var) 
numeric_vars_normal <- c("pH", "TA", "Nitrogen", "Humic_transformed", "Prot_transformed", "BIX", "HIX_transformed", "M_to_C")

# let's see all ANOVA results together
anovas_full <- map_dfr(numeric_vars_normal, function(v){
  df_sub <- testing_df_norm %>%
    filter(variable == v) # filter by variable
    options(contrasts = c("contr.sum", "contr.poly"))
    fit <- lmer(Value ~ Day_Night * Benthos * Season + (1| Quad_ID), data = df_sub)
    Anova(fit, type = "III") %>%
    as.data.frame() %>%
    rownames_to_column("term") %>%
    filter(term != "(Intercept)") %>%
    filter(`Pr(>Chisq)` < 0.05) %>% # see what's significant
    mutate(variable = v)
})
anovas_full %>%
  kable()
```
Summary Table

|Term                     | Variable(s) sig         |
|-------------------------|:-----------------------:|
|Day_Night                | TA, Humic, Prot, HIX    |
|Benthos                  | TA, NH4, Prot, BIX, M:C |
|Season                   | TA, NH4, Prot, M:C      |
|Benthos:Season           | Humic                   |
|Day_Night:Benthos        | PO4                     |
|Day_Night:Season         | PO4, NO2, HIX           |
|Day_Night:Benthos:Season | pH, TempInSitu, NO3, SIL|



Let's look at differences across benthos in their dawn/dusk values over the two seasons

# Heatmap colored in comparison to OO

```{r heatmap-plot-code}
# take means of all variables
heatmap_data <- data_clean %>%
  group_by(Benthos, Day_Night, Season) %>%
  summarise(across(all_of(numeric_vars), ~ mean(.x, na.rm = TRUE)),
            .groups = "drop") %>%
  pivot_longer(cols = all_of(numeric_vars),
               names_to = "variable",
               values_to = "Mean_Value")

# get open ocean means for comparison
open_ocean_ref <- heatmap_data %>%
  filter(Benthos == "Open Ocean") %>%
  select(-Benthos) %>%
  rename(OpenOcean_Value = Mean_Value)

# join both to calculate ΔOO
fdom_diff <- heatmap_data %>%
  filter(Benthos != "Open Ocean") %>% 
  left_join(open_ocean_ref,
            by = c("Day_Night","Season","variable")) %>%
  mutate(Diff_from_OO = Mean_Value - OpenOcean_Value)

# scale the differences so we can plot it
fdom_diff <- fdom_diff %>%
  group_by(variable) %>%
  mutate(Diff_Scaled = as.numeric(scale(Diff_from_OO))) %>%
  ungroup() %>%
  mutate(Benthos = factor(Benthos, levels = c("Barnacle", "Mussel", "Rockweed", "Surfgrass"))) %>%
  mutate(variable = factor(variable, levels =  c("pH", "TA", "Nitrogen", "Humic", "Prot", "BIX", "HIX", "M_to_C")))

# heat map colored by difference from open ocean
fdom_diff %>%
  ggplot(aes(x = variable,
             y = Day_Night,
             fill = Diff_Scaled)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#075AFF", mid = "white", high = "#FF4500",
    midpoint = 0, name = "Scaled difference\nfrom open ocean"
  ) +
  labs(
    title = "Deviation from Open Ocean fDOM + Nutrients",
    x = "Component",
    y = "Time of day"
  ) +
  theme_minimal() +
  facet_grid(Benthos ~ Season) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 10)
  )

```
## ANOVA
```{r heatmap-ANOVA, fig.height=10}
# filter out NAs
testing_df_norm <- testing_df_norm %>% 
  filter(!is.na(Value))%>%
  mutate(Season = as.factor(Season))

compute_sig_to_ctrl <- function(testing_df_norm, vars = all_numeric_vars_normal) {
  
  options(contrasts = c("contr.sum", "contr.poly")) # set contrasts for type III ANOVA
  
  res_tbl <- map(vars, function(v) {
    df <- testing_df_norm %>%
      filter(variable == v)
    
    fit <- lm(Value ~ Benthos * Day_Night * Season, data = df) # create lm
    model <- Anova(fit, type = "III")
    emm <- emmeans(fit, ~ Benthos | Day_Night * Season) # emmeans for tukey's hsd
    vs_ctrl <- contrast(emm, method = "trt.vs.ctrl", ref = control_label, adjust = "mvt")
    td  <- as.data.frame(vs_ctrl)
    
    comp_tbl <- td %>%
      separate(contrast, c("left","right"), sep = " - ", remove = FALSE) %>%
      mutate(
        Benthos  = if_else(right == control_label, left, right),
        variable = v,
        sig = case_when(
          is.na(p.value)  ~ NA_character_,
          p.value < 0.001 ~ "***",
          p.value < 0.01  ~ "**",
          p.value < 0.05  ~ "*",
          TRUE            ~ ""
        )
      ) 
    ctrl_tbl <- crossing(
      variable = v,
      Season   = unique(td$Season),
      Day_Night= unique(td$Day_Night),
      Benthos  = control_label
    ) %>%
      mutate(
        estimate = NA_real_, SE = NA_real_, df = NA_real_, t.ratio = NA_real_,
        p.value  = NA_real_, sig = ""
      )

    bind_rows(comp_tbl, ctrl_tbl) %>%
      distinct(variable, Season, Day_Night, Benthos, .keep_all = TRUE)
  }) %>% compact() %>% list_rbind()

  res_tbl
}

sig_to_ctrl <- compute_sig_to_ctrl(testing_df_norm) %>%
  mutate(variable = sub("_transformed$", "", variable)) # convert name back for join


# get n so we can add to the plot
n_table <- testing_df_norm %>%
  mutate(variable = sub("_transformed$", "", variable)) %>%
  group_by(Season, Benthos, variable, Day_Night) %>%
  summarise(n = sum(!is.na(Value)), .groups = "drop")

# join all data for plotting
plot_df <- fdom_diff %>%  
  left_join(sig_to_ctrl, by = c("Season","Day_Night","Benthos","variable")) %>%
  left_join(n_table,   by = c("Season","Day_Night","Benthos","variable"))

# create plot
plot_df %>%
  ggplot(aes(x = variable,
             y = Day_Night,
             fill = Diff_Scaled)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#075AFF", mid = "white", high = "#FF4500",
    midpoint = 0, name = "Scaled difference\nfrom open ocean"
  ) +
  labs(
    title = "Dusk vs Dawn fDOM + Nutrients (orig ANOVA)",
    x = "Component",
    y = "Time of day"
  ) +
  geom_text(aes(label = ifelse(is.na(sig),"",paste0(sig,"\nn\n",n))), color = "black", size = 3) +
  theme_minimal() +
  facet_grid(Benthos ~ Season) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 10)
  )
```

## One-sample t-test
```{r heatmap-ttest, fig.height = 10}
compute_sig_to_ctrl_ttest <- function(testing_df_norm, 
                                      vars = all_numeric_vars_normal,
                                      control_label) {
  
  res_tbl <- map(vars, function(v) { # loop for each variable
    df_var <- testing_df_norm %>%
      filter(variable == v)
    
    res_by_sd <- df_var %>%
      group_nest(Season, Day_Night) %>% # group by season and dawn/dusk
      mutate(
        res = map(data, function(df_sd) {
          ctrl_vals <- df_sd %>%
            filter(Benthos == control_label) %>%
            pull(Value)
          ctrl_mean <- mean(ctrl_vals, na.rm = TRUE) # get OO mean

          df_sd %>%
            group_by(Benthos) %>%
            summarise(
              n       = sum(!is.na(Value)),
              mean    = mean(Value, na.rm = TRUE),
              .groups = "drop"
            ) %>%
            mutate(
              # one-sample test vs control mean for each benthos
              p.value = map_dbl(Benthos, function(b) {
                x <- df_sd %>% filter(Benthos == b) %>% pull(Value)
                t.test(x, mu = ctrl_mean)$p.value
              }),
              estimate = mean - ctrl_mean,
              variable = v
            )
        })
      ) %>%
      select(-data) %>%
      unnest(res)
    
    # add sig stars
    res_by_sd %>%
      mutate(
        sig = case_when(
          is.na(p.value)  ~ NA_character_,
          p.value < 0.001 ~ "***",
          p.value < 0.01  ~ "**",
          p.value < 0.05  ~ "*",
          TRUE            ~ ""
        )
      )
  }) %>% 
    compact() %>% 
    list_rbind()
  
  res_tbl
}

sig_to_ctrl_ttest <- compute_sig_to_ctrl_ttest(testing_df_norm,
                                               vars = all_numeric_vars_normal,
                                               control_label = control_label) %>%
  mutate(variable = sub("_transformed$", "", variable))

plot_df_ttest <- fdom_diff %>%
  left_join(sig_to_ctrl_ttest, 
            by = c("Season","Day_Night","Benthos","variable"))

# create plot
plot_df_ttest %>%
  ggplot(aes(x = variable,
             y = Day_Night,
             fill = Diff_Scaled)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#075AFF", mid = "white", high = "#FF4500",
    midpoint = 0, name = "Scaled difference\nfrom open ocean"
  ) +
  labs(
    title = "Dusk vs Dawn fDOM + Nutrients (t-test vs control)",
    x = "Component",
    y = "Time of day"
  ) +
  geom_text(aes(label = ifelse(is.na(sig), "", paste0(sig, "\nn\n", n))),
            color = "black", size = 3) +
  theme_minimal() +
  facet_grid(Benthos ~ Season) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text  = element_text(size = 10)
  )

```

## Dunn's test
```{r heatmap-dunn, fig.height = 10}
compute_sig_to_ctrl_dunn <- function(testing_df_norm, 
                                     vars = numeric_vars,
                                     control_label) {
  
  res_tbl <- map(vars, function(v) { # loop for each variable
    df_var <- testing_df_norm %>%
      filter(variable == v)
    
    df_var %>%
      group_by(Season, Day_Night) %>% # group by season and dawn/dusk
      group_modify(~ {
        df_sd <- .x
  
        dunn_res <- df_sd %>%
          rstatix::dunn_test(Value ~ Benthos, p.adjust.method = "bonferroni")
        
        # keep only comparisons involving OO
        dunn_ctrl <- dunn_res %>%
          filter(group1 == control_label | group2 == control_label) %>%
          mutate(
            Benthos  = if_else(group1 == control_label, group2, group1),
            variable = v,
            p.value  = p.adj
          )
        
        dunn_ctrl
      }) %>%
      ungroup()
  }) %>%
    compact() %>%
    list_rbind() %>%
    mutate(
      sig = case_when(
        is.na(p.value)  ~ NA_character_,
        p.value < 0.001 ~ "***",
        p.value < 0.01  ~ "**",
        p.value < 0.05  ~ "*",
        TRUE            ~ ""
      )
    )
  
  res_tbl
}

sig_to_ctrl_dunn <- compute_sig_to_ctrl_dunn(testing_df_norm,
                                             vars = numeric_vars,
                                             control_label = control_label) %>%
  mutate(variable = sub("_transformed$", "", variable))

plot_df_dunn <- fdom_diff %>%
  left_join(sig_to_ctrl_dunn, 
            by = c("Season","Day_Night","Benthos","variable")) %>%
  left_join(n_table, 
            by = c("Season","Day_Night","Benthos","variable"))

# create plot
plot_df_dunn %>%
  ggplot(aes(x = variable,
             y = Day_Night,
             fill = Diff_Scaled)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#075AFF", mid = "white", high = "#FF4500",
    midpoint = 0, name = "Scaled difference\nfrom open ocean"
  ) +
  labs(
    title = "Dusk vs Dawn fDOM + Nutrients (Dunn’s test vs control)",
    x = "Component",
    y = "Time of day"
  ) +
  geom_text(aes(label = ifelse(is.na(sig), "", paste0(sig, "\nn\n", n))),
            color = "black", size = 3) +
  theme_minimal() +
  facet_grid(Benthos ~ Season) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text  = element_text(size = 10)
  )

```

# Create effect size plots

```{r effect-size-plot}

build_diel_effect_df_one_scaled <- function(var_name, data = data_clean) {
  stopifnot(var_name %in% names(data))
  
  # scale the variable (global z-score)
  data_z <- data %>%
    mutate(var_z = as.numeric(scale(.data[[var_name]])))
  
  fml <- as.formula("var_z ~ Day_Night * Season * Benthos + (1|Quad_ID)")
  fit <- lmer(fml, data = data_z)
  
  # emmeans for Day_Night within each Season × Benthos
  emm <- emmeans(fit, ~ Day_Night | Season * Benthos)
  
  # Dusk - Dawn contrast in *scaled units* (≈ SD units)
  diel <- contrast(emm, method = "revpairwise")
  
  # summary with p-values + 95% CI
  df <- summary(diel, infer = c(TRUE, TRUE), level = 0.95) |>
    as.data.frame()
  
  # CI column names can vary; harmonize to lower.CL / upper.CL
  if (!"lower.CL" %in% names(df)) {
    if ("asymp.LCL" %in% names(df)) {
      df <- df |>
        rename(
          lower.CL = asymp.LCL,
          upper.CL = asymp.UCL
        )
    } else if ("LCL" %in% names(df)) {
      df <- df |>
        rename(
          lower.CL = LCL,
          upper.CL = UCL
        )
    }
  }
  
  df %>%
    mutate(
      est_std   = estimate,     
      lower_std = lower.CL,
      upper_std = upper.CL,
      variable  = var_name
    )
}

effects_all <- map_dfr(
  numeric_vars,
  build_diel_effect_df_one_scaled,
  data = data_clean
)

effects_all <- effects_all |>
  mutate(
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ "ns"
    ),
    variable = factor(variable, levels = rev(numeric_vars))
  )

ggplot(effects_all,
       aes(x = est_std, y = variable,
           color = Season,  
           alpha = sig != "ns")) +   
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey70") +
  geom_errorbarh(aes(xmin = lower_std, xmax = upper_std),
                 height = 0.15) +
  geom_point(size = 2) +
  facet_wrap(~ Benthos, ncol = 5) +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.3),
                     guide = "none") +
  scale_color_manual(values = c(
    "Fall"   = "#D95F02",
    "Spring" = "#1B9E77"
  )) +
  labs(
    title = "Standardized Dusk–Dawn Effect Sizes by Variable and Benthos",
    x = "Standardized effect size (Dusk – Dawn, SD units)",
    y = NULL,
    color = "Season"
  ) +
  
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey90"),
    axis.text.y = element_text(size = 9)
  )

```

# Diel boxplots with significance levels

```{r diel-boxplots}
build_sig_df <- function(var_name, data = data_clean) {
  stopifnot(var_name %in% names(data))
  
  # fit ANOVA
  fml <- as.formula(paste(var_name, "~ Day_Night * Season * Benthos + (1|Quad_ID)"))
  fit <- lmer(fml, data = data)
  
  # Dawn vs Dusk within each Season × Benthos
  emm  <- emmeans(fit, ~ Day_Night | Season * Benthos)
  contr_df <- as.data.frame(contrast(emm, method = "revpairwise"))
  # columns: Season, Benthos, contrast, estimate, SE, df, t.ratio, p.value
  
  # Turn p-values into stars
  sig_df <- contr_df |>
    select(Season, Benthos, p.value) |>
    distinct() |>
    mutate(sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ "ns"
    )) |>
    left_join(
      data |>
        group_by(Season, Benthos) |>
        summarise(y_pos = max(.data[[var_name]], na.rm = TRUE),
                  .groups = "drop"),
      by = c("Season", "Benthos")
    ) |>
    mutate(y_pos = y_pos * 1.05)   # a bit above top whisker
  
  # drop non-significant labels to reduce clutter
  sig_df |> filter(sig != "ns")
}

plot_diel_with_sig <- function(var_name, var_label, 
                               data = data_clean,
                               free_y = FALSE) {
  stopifnot(var_name %in% names(data))
  
  sig_df <- build_sig_df(var_name, data = data)
  
  facet_scales <- if (free_y) "free_y" else "fixed"
  
  ggplot(
    data |> filter(!is.na(.data[[var_name]])),
    aes(x = Day_Night, y = .data[[var_name]], fill = Day_Night)
  ) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
    facet_wrap(Season ~ Benthos, ncol = 5, scales = facet_scales) +
    scale_fill_manual(values = c("Dawn" = "#FDB813", "Dusk" = "#003865")) +
    labs(
      title = paste("Diel Patterns in", var_label),
      x = "Time of Day",
      y = var_label
    ) +
    theme(
      legend.position = "none",
      strip.background = element_rect(fill = "grey90"),
      axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
    geom_text(
      data = sig_df,
      aes(x = 1.5, y = y_pos, label = sig),
      inherit.aes = FALSE,
      size = 4
    )
}

```

## Example plots

```{r diel-boxplot-ex}
(p_ph   <- plot_diel_with_sig("pH",   "pH (Total Scale)"))
(p_no3  <- plot_diel_with_sig("NO3",  "Nitrate (µmol/L)", free_y = TRUE))
(p_sil  <- plot_diel_with_sig("SIL",  "Silicate (µmol/L)"))
(p_po4  <- plot_diel_with_sig("PO4",  "Phosphate (µmol/L)"))
(p_TA   <- plot_diel_with_sig("TA",   "Total Alkalinity (µmol/kg)"))
(p_hix  <- plot_diel_with_sig("HIX",  "HIX"))
(p_hum  <- plot_diel_with_sig("Humic","Total Humic-like fDOM (RFU)"))
(p_prot <- plot_diel_with_sig("Prot", "Total Proteinaceous-like fDOM (RFU)"))
(p_mc <- plot_diel_with_sig("M_to_C", "M:C"))
```

# Community Composition

## Sessile
```{r sessile-barplot}
## sum across the canopy and understory ###

sessile_total <-sessile %>%
  group_by(plot, season)%>%
  summarise_at(.vars = vars(Silvetia_compressa:Chondracanthus_canaliculatus), .funs = sum)%>% 
  mutate(Total_hairs = rowSums(across(Silvetia_compressa:Chondracanthus_canaliculatus))) %>% # count total crosshairs
  ungroup()%>% # calculate percent cover
  mutate(across(c(Silvetia_compressa:Chondracanthus_canaliculatus),  ~ . /Total_hairs*100))%>%
  rename(BenthicID = plot, Season = season)%>% # keep consistent with other data
  mutate(Season = factor(Season, levels = c("summer2023","fall2023","winter2024", "spring2024")))%>%
  mutate(Benthos = case_when( # add benthos name
    grepl("B", BenthicID) ~ "Barnacles",
    grepl("M", BenthicID) ~"Mussels",
    grepl("P", BenthicID) ~"Surfgrass",
    grepl("R", BenthicID) ~"Rockweed"), 
    .after = BenthicID)


## Make a stacked barplot of all the species across seasons
sessile_total %>%
  group_by(Benthos, Season)%>% # calculate the average for each season
  summarise(across(c(Silvetia_compressa:Chondracanthus_canaliculatus),  ~mean(.,na.rm = TRUE))) %>%
  pivot_longer(cols = c(Silvetia_compressa:Chondracanthus_canaliculatus), names_to = "Species", values_to = "Percent_Cover")%>%
  left_join(sessile_function)%>%
  filter(Season %in% c('fall2023','spring2024')) %>%
  ggplot(aes(x = Season,y = Percent_Cover, fill = FunctionalGroup ))+
    geom_bar(position="stack", stat="identity")+
  scale_fill_paletteer_d("MetBrewer::Signac")+
  labs(x = "",
       y = "% Cover",
       color = "")+
  theme_bw()+
  facet_wrap(~Benthos)

#ggsave(here("Output","SessileComp.png"), width = 10, height = 6)
```

## Mobiles
```{r mobile-barplot}

## calculate denisity per m2 (used 0.25m2 quad)
mobile_density <-
  mobile %>%
  rename(BenthicID = plot, Season = season)%>% # keep consistent with other data
  mutate(Season = factor(Season, levels = c("summer2023","fall2023","winter2024","spring2024")))%>%
  mutate(Benthos = case_when( # add benthos name
    grepl("B", BenthicID) ~ "Barnacles",
    grepl("M", BenthicID) ~"Mussels",
    grepl("P", BenthicID) ~"Surfgrass",
    grepl("R", BenthicID) ~"Rockweed"), 
    .after = BenthicID)%>%
  select(-c(Littorina_sp_raw, Lottia_sp_raw))%>%
  mutate(across(c(Lottia_sp_scaled:Strongylocentrotus_purpuratus),  ~./0.25))
  

mobile_density %>%
  group_by(Benthos, Season)%>% # calculate the average for each season
  summarise(across(c(Lottia_sp_scaled:Strongylocentrotus_purpuratus),  ~mean(.,na.rm = TRUE))) %>%
  pivot_longer(cols = c(Lottia_sp_scaled:Strongylocentrotus_purpuratus), names_to = "Species", values_to = "Density_m2")%>%
  left_join(mobile_function)%>% #join with functional data
  filter(Season %in% c('fall2023','spring2024')) %>%
  ggplot(aes(x = Season,y = Density_m2, fill = FunctionalGroup ))+
  geom_bar(position="stack", stat="identity")+
  scale_fill_paletteer_d("MetBrewer::Signac")+
  labs(x = "",
       y = "Mean Density (m2)",
       color = "")+
  theme_bw()+
  facet_wrap(~Benthos, scale = "free")

#ggsave(here("Output","MobileComp.png"), width = 10, height = 6)


## rename the foundation species data to match the chemistry
sessile_clean<-sessile_total %>%
  rename(Quad_ID = BenthicID) %>%
  mutate(Season = case_when(Season  == "fall2023"~"Fall",
                            Season == "spring2024"~"Spring",
                            Season == "summer2023"~"Summer",
                            Season == "winter2024"~"Winter")) %>%
  mutate(FoundationSp = case_when(Benthos == "Barnacles"~ `Balanus/Cthamalus_sp`,
                                   Benthos == "Rockweed"~Silvetia_compressa,
                                   Benthos == "Surfgrass"~Phyllospadix_sp,
                                   Benthos == "Mussels"~Mytilus_californianus)) %>%  # add a column that is the cover of the foundation species for the right benthos
  select(!Total_hairs) %>%
  mutate(Benthos = ifelse(Benthos == "Mussels","Mussel",Benthos))%>%
  mutate(Benthos = ifelse(Benthos == "Barnacles","Barnacle",Benthos))

#write_csv(x = sessile_clean, here("Data","Community_Composition","Sessile_clean.csv"))

```